---
title: "Generalised Additive Model"
author: "Denisse Fierro Arcos"
date: "2023-09-22"
output: 
  github_document:
    toc: true
    html_preview: false
---

# Generalised Additive Model (GAM)

GAMs are commonly used in species distribution modelling because of their flexibility. GAMs use smoothing functions to capture non-linear relationships between the dependent variable and the predictors (i.e., independent variables).

In this project, we will use GAMs as one of the models to be considered in our Species Distribution Model ensemble to estimate the distribution of crabeater seals in the recent past.  
  
## Loading libraries

```{r, message=F, warning=F}
library(tidyverse)
library(mgcv)
library(stars)
library(cmocean)
library(cowplot)
source("useful_functions.R")
```


## Setting up notebook

Selecting an output folder for GAM results exists and getting a list of data files.  
  
```{r}
#Location of folder for outputs
out_folder <- "../../SDM_outputs/GAM"
#If folder does not exist, create one
if(!dir.exists(out_folder)){
  dir.create(out_folder, recursive = T)
}

#Get path to files containing data
file_list <- list.files("../../Environmental_Data/", pattern = "Indian", full.names = T)
```
  
### Loading mean environmental conditions from ACCESS-OM2-01

This dataset includes the mean environmental conditions per month (November and December) over the entire period of study (1981 to 2013).

```{r}
mean_model <- read_csv("../../Environmental_Data/ACCESS-OM2-01/All_values_month_ACCESS-OM2-01_env_vars.csv") %>% 
  mutate(month = factor(month)) %>% 
  #Drop variables with high multicollinearity
  select(!c(freez_pot_Wm2, bottom_sal_psu, SIT_m))

#List of categorical variables
cat_vars <- c("month")

mean_model_baked <- prep_pred(mean_model, cat_vars)
```
  
### Loading mean environmental conditions from observations

This dataset includes the mean environmental conditions per month (November and December) over the entire period of study (1981 to 2013).

```{r}
mean_obs <- read_csv("../../Environmental_Data/Env_obs/All_values_month_Obs_env_vars.csv") %>% 
  mutate(month = factor(month))

#List of categorical variables
cat_vars <- c("month")

mean_obs_baked <- prep_pred(mean_obs, cat_vars)
```
  
## Loading layers for plotting
We will extract this layer from the `rnaturalearth` package. We will then reproject this layer to South Polar Stereographic (`EPSG 3976`).  
  
```{r}
#Loading layer
antarctica <- rnaturalearth::ne_countries(continent = "Antarctica",
                                          returnclass = "sf") %>% 
  #Transforming to South Polar Stereographic
  st_transform(3976)
```


## Loading environmental data from ACCESS-OM2-01 and setting up variables

We will use the datasets created in the notebook `02_Merging_background_presence_data.Rmd` located within the `Scripts/05_SDMs` folder. These datasets include the crabeater seal observations, background points, and environmental data.  
  
We will also define categorical and continuous explanatory variables.  
  
## Environmental variables matching observations 
First, we will look only at the variables with no multicollinearity. This means that sea surface temperature (`SST`) is excluded even though this variable is available in the observational dataset.  
  
The variable `month` will be included as an ordinal factor in our analysis.  
  
```{r}
#Loading data
mod_match_obs <- read_csv(str_subset(file_list, "match")) %>% 
  #Setting month as factor and ordered factor
  mutate(month = factor(month))

#List of covariates
covars <- str_subset(names(mod_match_obs), "presence|_ocean", negate = T)
```
  
### Building GAM formula
We will run an initial GAM using all available covariates. Sea ice related variables will be fitted individually to each `month` as we expect sea ice to vary in extent between November and December because sea ice should be retreating during this time.  
  
```{r}
# Most complex model
full_model <- presence ~ month + s(year) + s(bottom_slope_deg) + s(dist_shelf_km) + s(dist_coast_km) + s(depth_m) + s(SIC, by = month) + s(lt_pack_ice, by = month) + s(dist_ice_edge_km, by = month)
```
    
### Splitting data into testing and training
The `prep_data` function in the `useful_functions` script will be used to split our data and to apply all necessary transformations.  
  
```{r}
#Getting training data
mod_match_obs_split <- prep_data(mod_match_obs, cat_vars)

#Selecting model data
model_data <- mod_match_obs_split$baked_train %>%
  select(all_of(covars) | "presence")
```
    
### Modelling
Background data (`presence` == `0`) will be down-weighted because they do not truly represent absences. The down-weighting applied represents the proportion of presence to background points. In this way, the sum of the weighted background points will be the same as the total of presences.  
  
We will run the GAM using all covariates available and we will downweight background points so they have the same effect as all presences.  
  
```{r}
#Calculating downweights
weights <- down_weights(model_data)

#Full model
full_model_gam <- gam(formula = as.formula(full_model),
    data = model_data,
    family = binomial(link = "cloglog"),
    weights = weights,
    method = "REML", select = T)

#Print summary
summary(full_model_gam)
```
  
Removing variables with effects reduced to near zero (`edf` < 0.001) or that are non-significant: `year`, `dist_shelf_km` (distance to continental shelf), `bottom_slope_deg` (seafloor slope), and `dist_coast_km` (distance to coastline in km).
    
```{r}
#Defining new formula without non-significant covariates
significant_only <-  presence ~ month + s(depth_m) + s(SIC, by = month) + s(lt_pack_ice, by = month) + s(dist_ice_edge_km, by = month)

#Applying new formula
significant_only_gam <- gam(formula = as.formula(significant_only),
    data = model_data,
    family = binomial(link = "cloglog"),
    weights = weights,
    method = "REML", select = T)

summary(significant_only_gam)
```
  
### Comparing models
The Akaike's Information Criterion (AIC) will be calculated to assess model performance. We will also check the $r^{2}$ values for both models.    

```{r}
#Putting everything in a data frame
sum_gam <- data.frame(model = c("full_model", "significant_only"),
                      AIC = c(AIC(full_model_gam), AIC(significant_only_gam)),
                      rsq = c(summary(full_model_gam)$r.sq, summary(significant_only_gam)$r.sq))
#Checking results
sum_gam
```
  
These models performed almost the same, so we will keep the most parsimonious model (i.e., `significant_only`), which has a slightly lower AIC and slightly higher $r^{2}$.  We will include `SST` to this model and check the effect of `SST` on model performance. The smoothing for this variable will vary by month as we have done with sea ice related variables.  
  
```{r}
significant_only_SST <-  presence ~ month + s(depth_m) + s(SIC, by = month) + s(lt_pack_ice, by = month) + s(dist_ice_edge_km, by = month) + s(SST_degC, by = month)

#Applying new formula
significant_only_SST_gam <- gam(formula = as.formula(significant_only_SST),
    data = model_data,
    family = binomial(link = "cloglog"),
    weights = weights,
    method = "REML", select = T)

#Check results
summary(significant_only_SST_gam)
```
  
The model that includes `SST` explains almost 5\% of variability in the data, which is higher than the previous models. We can see that now sea ice concentration (`SIC`) for November has become insignificant, and this is likely because of its high correlation to `SST`. We will remove `SIC`, and run the model once more to see its effect on deviance explained.  
  
```{r}
significant_only_no_SIC <-  presence ~ month + s(depth_m) + s(lt_pack_ice, by = month) + s(dist_ice_edge_km, by = month) + s(SST_degC, by = month)

#Applying new formula
significant_only_no_SIC_gam <- gam(formula = as.formula(significant_only_no_SIC),
    data = model_data,
    family = binomial(link = "cloglog"),
    weights = weights,
    method = "REML", select = T)

#Check results
summary(significant_only_no_SIC_gam)
```
  
We will now calculate AIC for the last two models we ran, extract their $r^{2}$ values and add them to our summary table.  
  
```{r}
#Adding results to summary table
sum_gam <- data.frame(model = c("significant_only_SST", "significant_only_no_SIC"),
                      AIC = c(AIC(significant_only_SST_gam), AIC(significant_only_no_SIC_gam)),
                      rsq = c(summary(significant_only_SST_gam)$r.sq, summary(significant_only_no_SIC_gam)$r.sq)) %>% 
  bind_rows(sum_gam, .) %>% 
  #Arrange in increasing order by AIC
  arrange(AIC)

#Checking results
sum_gam
```
  
We can see that the best performing model based on AIC is the last one we tested, which includes `SST`, and `SIC`. We can check if the performance of this model differs significantly from the model that includes `SST`, but excludes `SIC` (`significant_only_no_SIC` in table above). To do this, we will performance an ANOVA test.  
  
```{r}
anova(significant_only_SST_gam, significant_only_no_SIC_gam, test = "Chisq")
```
  
We can see these models differ significantly, but since multicollinearity has a negative effect on GAM predicitive ability, we will use the `significant_only_no_SIC` to predict crabeater distribution.  
  
### Predictions
We will use the best performing model to predict crabeater seal distribution using mean monthly environmental conditions obtained from ACCESS-OM2-01.  
  
```{r}
#Prediction monthly crabeater seal distribution
mean_pred_match_obs <- mean_model_baked %>% 
  mutate(pred = as.vector(predict(significant_only_no_SIC_gam, 
                                  mean_model_baked, 
                                  type = "response")))

#Converting from data frame to raster
mean_pred_match_obs_ras <- mean_pred_match_obs %>%
  #Select relevant variables only
  select(xt_ocean, yt_ocean, pred, month) %>% 
  #Set dimensions
  st_as_stars(dims = c("xt_ocean", "yt_ocean", "month")) %>% 
  #Ensuring month dimension is shown correctly
  st_set_dimensions("month", values = c(11, 12)) %>%
  #Set CRS
  st_set_crs(4326) %>% 
  #Transform to South Pole stereographic
  st_transform(crs = st_crs(3976))

#Saving outputs
#Data frame
mean_pred_match_obs %>% 
  write_csv(file.path(out_folder, "mean_pred_match_obs.csv"))
#Saving as R dataset so it can be easily open with readRDS
saveRDS(mean_pred_match_obs_ras,
        file.path(out_folder, "mean_pred_match_obs_raster.rds"))
```
  
Now that we have predicted the mean distribution of crabeaters based with our select model, we can plot results for comparison later.  
  
```{r}
#Plotting November distribution
#Prepping data
nov <- mean_pred_match_obs_ras %>% 
  slice(index = 1, along = "month") 

#Plotting
nov_plot <- ggplot()+
  geom_stars(data = nov)+
  geom_sf(data = antarctica)+
  lims(x = c(0, 4000000))+
  #Set colour palette
  scale_fill_cmocean(name = "haline", direction = -1, 
                     guide = guide_colorbar(barwidth = 1, barheight = 10, 
                                            ticks = FALSE, nbin = 1000, 
                                            frame.colour = "black"), 
                     limits = c(0, 1)) +
  theme_linedraw() +
  theme(panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), 
        legend.position = "none",
        plot.title = element_text(hjust = 0.5)) +
  labs(title = "November",
       x = "Longitude",
       y = "Latitude")

dec <- mean_pred_match_obs_ras %>% 
  slice(index = 2, along = "month") 

dec_plot <- ggplot() +
  geom_stars(data = dec) +
  geom_sf(data = antarctica)+
  lims(x = c(0, 4000000))+
  scale_fill_cmocean(name = "haline", direction = -1, 
                     guide = guide_colorbar(barwidth = 1, barheight = 10, 
                                            ticks = FALSE, nbin = 1000, 
                                            frame.colour = "black"), 
                     limits = c(0, 1)) +
  theme_linedraw() +
  theme(panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), 
        plot.title = element_text(hjust = 0.5)) +
  labs(title = "December",
       x = "Longitude",
       y = " ",
       fill = "Probability")

#Get legend
legend <- get_legend(dec_plot)

#Remove legend from December plot
dec_plot <- dec_plot + theme(legend.position = 'none')

#Plotting together
plot_match_obs <- plot_grid(nov_plot, dec_plot, legend, ncol = 3, nrow = 1,
                            rel_widths = c(1, 1, 0.3))

#Add title
title <- ggdraw()+
  draw_label("Mean crabeater seal distribution\n(ACCESS-OM2-01 - simplified)",
             fontface = "bold", hjust = 0.5)+
  theme(plot.margin = margin(0, 0, 0, 0))

#Putting everything together
final <- plot_grid(title, plot_match_obs, ncol = 1, 
                   rel_heights = c(0.1, 1))

#Saving graph
ggsave(file.path(out_folder, "map_mean_pred_match_obs.png"), 
       plot = final, device = "png", bg = "white", width = 8.75, height = 7)
```
  
## All environmental variables available in ACCESS-OM2-01
We will now use the full range of covariates available in the ACCESS-OM2-01 model. We will test if the inclusion of additional environmental variables improve predictive ability. We will follow the same approach as we did before, first we will test a model with all covariates and exclude any with non-significant effects.  
  
We tested this dataset for multicollinearity, so we will get the names of variables with low VIF plus SST as this was found to be an important covariate before.  
  
```{r}
#Loading data
full_mod <- read_csv(str_subset(file_list, "model_env_pres")) %>% 
  #Setting month as factor and ordered factor
  mutate(month = factor(month))

#List of variables to be extracted from main dataset
mod_vars <- read_csv("../../Environmental_Data/ACCESS-OM2-01/Obs_BG_5x_Indian_weaning_LowVIF.csv") %>% 
  select(!c(sector, zone, season_year:decade)) %>% 
  names() %>% 
  append(c("SST_degC"))

#Extracting relevant columns only
full_mod <- full_mod %>% 
  select(all_of(mod_vars))

#List of environmental covariates to be used in model
covars <- str_subset(names(full_mod), "presence|_ocean", negate = T)
```
  
### Building GAM formula
We will run an initial GAM using all available covariates. Non-static variables (i.e., variables that vary with time) will be fitted individually to each `month` as we expect them to vary between November and December as we move from spring to summer.  
  
```{r}
# Most complex model
full_model <- presence ~ month + s(year) + s(bottom_slope_deg) + s(dist_shelf_km) + s(dist_coast_km) + s(depth_m) + s(SIC, by = month) + s(bottom_temp_degC, by = month) + s(SSS_psu, by = month) + s(vel_lat_surf_msec, by = month) + s(vel_lat_bottom_msec, by = month) + s(vel_lon_surf_msec, by = month) + s(vel_lon_bottom_msec, by = month) + s(lt_pack_ice, by = month) + s(dist_ice_edge_km, by = month)
```
    
### Splitting data into testing and training
The `prep_data` function in the `useful_functions` script will be used to split our data and to apply all necessary transformations.  
  
```{r}
#Getting training data
full_mod_split <- prep_data(full_mod, cat_vars)

#Selecting model data
model_data <- full_mod_split$baked_train %>%
  select(all_of(covars) | "presence")
```
  
### Modelling
Background data will be down-weighted and then we will run the GAM using all covariates available.  
  
```{r}
#Calculating downweights
weights <- down_weights(model_data)

#Full model
full_model_gam <- gam(formula = as.formula(full_model),
    data = model_data,
    family = binomial(link = "cloglog"),
    weights = weights,
    method = "REML", select = T)

#Print summary
summary(full_model_gam)
```  
  
This model explains more of the variability in our data than the model that uses variables for which we also have observations. We will remove the variables with a low contribution (`edf` $\sim 0.001$). This includes `year`, bottom slope (`bottom_slope_deg`), distance to the coastline (`dist_coast_km`), and bottom meridional water velocity (`vel_lon_bottom_msec`).

We will defined this simplified model and test it below.  
  
```{r}
# Simplified model
simpler_model <- presence ~ month + s(dist_shelf_km) + s(depth_m) + s(SIC, by = month) + s(bottom_temp_degC, by = month) + s(SSS_psu, by = month) + s(vel_lat_surf_msec, by = month) + s(vel_lat_bottom_msec, by = month) + s(vel_lon_surf_msec, by = month) + s(lt_pack_ice, by = month) + s(dist_ice_edge_km, by = month)

#Applying new simplified formula
simpler_model_gam <- gam(formula = as.formula(simpler_model),
    data = model_data,
    family = binomial(link = "cloglog"),
    weights = weights,
    method = "REML", select = T)

#Print summary
summary(simpler_model_gam)
```
  
We can see that this simplified model performs almost the same as the model including all environmental variables available.  
  
We will now explore the influence of `SST` as it was identified as an influential variable in the previous dataset.  
  
```{r}
# Simplified model with SST
simpler_model_SST <- presence ~ month + s(dist_shelf_km) + s(depth_m) + s(SIC, by = month) + s(bottom_temp_degC, by = month) + s(SSS_psu, by = month) + s(vel_lat_surf_msec, by = month) + s(vel_lat_bottom_msec, by = month) + s(vel_lon_surf_msec, by = month) + s(lt_pack_ice, by = month) + s(dist_ice_edge_km, by = month) + s(SST_degC, by = month)

#Applying new simplified formula
simpler_model_SST_gam <- gam(formula = as.formula(simpler_model_SST),
    data = model_data,
    family = binomial(link = "cloglog"),
    weights = weights,
    method = "REML", select = T)

#Print summary
summary(simpler_model_SST_gam)
```
  
This model is slightly worse than when `SST` was not included. This is likely because `SST` was highly correlated to `SIC`. Therefore, we will not use `SST` to predict crabeater seal distribution.  
  
Now that `SST` has been included, salinity at the surface (`SSS_psu`) has become non-significant. The model predictive ability increased, but since we know `SIC` is highly correlated to `SST`, we will check the effect that removing `SIC` has on predictive ability.  
  
```{r}
# Simplified model with SST and no SIC
simpler_model_SST_noSIC <- presence ~ month + s(dist_shelf_km) + s(depth_m) + s(bottom_temp_degC, by = month) + s(SSS_psu, by = month) + s(vel_lat_surf_msec, by = month) + s(vel_lat_bottom_msec, by = month) + s(vel_lon_surf_msec, by = month) + s(lt_pack_ice, by = month) + s(dist_ice_edge_km, by = month) + s(SST_degC, by = month)

#Applying new simplified formula
simpler_model_SST_noSIC_gam <- gam(formula = as.formula(simpler_model_SST_noSIC),
    data = model_data,
    family = binomial(link = "cloglog"),
    weights = weights,
    method = "REML", select = T)

#Print summary
summary(simpler_model_SST_noSIC_gam)
```
  
### Comparing models
Once again, we will calculate AIC for each model and obtain the coefficient of determination ($r^{2}$) to assess model performance.   

```{r}
full_sum_gam <- data.frame(model = c("full_model", "simpler_model", "simpler_model_SST", "simpler_model_SST_noSIC"),
                      AIC = c(AIC(full_model_gam), AIC(simpler_model_gam), AIC(simpler_model_SST_gam),
                              AIC(simpler_model_SST_noSIC_gam)),
                      rsq = c(summary(full_model_gam)$r.sq, summary(simpler_model_gam)$r.sq,
                              summary(simpler_model_SST_gam)$r.sq, summary(simpler_model_SST_noSIC_gam)$r.sq))

#Arranging models from lowest to highest AIC
full_sum_gam %>% 
  arrange(AIC)
```
  
Based on AIC and $r^{2}$, we can see that the models including `SST` are the best performing. The exclusion of `SIC` results in a slightly lower predictive ability, but we will use this latter model for predictions as we know GAMs are affected by multicollinearity.  
  
### Predictions
We will use the best performing model to predict crabeater seal distribution using mean monthly environmental conditions obtained from ACCESS-OM2-01.  
  
```{r}
#Prediction monthly crabeater seal distribution
mean_pred_mod <- mean_model_baked %>% 
  mutate(pred = as.vector(predict(simpler_model_SST_noSIC_gam, 
                                  mean_model_baked, 
                                  type = "response")))

#Converting from data frame to raster
mean_pred_mod_ras <- mean_pred_mod %>%
  #Select relevant variables only
  select(xt_ocean, yt_ocean, pred, month) %>% 
  #Set dimensions
  st_as_stars(dims = c("xt_ocean", "yt_ocean", "month")) %>% 
  #Ensuring month dimension is shown correctly
  st_set_dimensions("month", values = c(11, 12)) %>%
  #Set CRS
  st_set_crs(4326) %>% 
  #Transform to South Pole stereographic
  st_transform(crs = st_crs(3976))

#Saving outputs
#Data frame
mean_pred_mod %>% 
  write_csv(file.path(out_folder, "mean_pred_mod.csv"))
#Saving as R dataset so it can be easily open with readRDS
saveRDS(mean_pred_mod_ras,
        file.path(out_folder, "mean_pred_mod_raster.rds"))
```
  
Now that we have predicted the mean distribution of crabeaters based with our select model, we can plot results for comparison later.  
  
```{r}
#Plotting November distribution
#Prepping data
nov <- mean_pred_mod_ras %>% 
  slice(index = 1, along = "month") 

#Plotting
nov_plot <- ggplot()+
  geom_stars(data = nov)+
  geom_sf(data = antarctica)+
  lims(x = c(0, 4000000))+
  #Set colour palette
  scale_fill_cmocean(name = "haline", direction = -1, 
                     guide = guide_colorbar(barwidth = 1, barheight = 10, 
                                            ticks = FALSE, nbin = 1000, 
                                            frame.colour = "black"), 
                     limits = c(0, 1)) +
  theme_linedraw() +
  theme(panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), 
        legend.position = "none",
        plot.title = element_text(hjust = 0.5)) +
  labs(title = "November",
       x = "Longitude",
       y = "Latitude")

dec <- mean_pred_mod_ras %>% 
  slice(index = 2, along = "month") 

dec_plot <- ggplot() +
  geom_stars(data = dec) +
  geom_sf(data = antarctica)+
  lims(x = c(0, 4000000))+
  scale_fill_cmocean(name = "haline", direction = -1, 
                     guide = guide_colorbar(barwidth = 1, barheight = 10, 
                                            ticks = FALSE, nbin = 1000, 
                                            frame.colour = "black"), 
                     limits = c(0, 1)) +
  theme_linedraw() +
  theme(panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), 
        plot.title = element_text(hjust = 0.5)) +
  labs(title = "December",
       x = "Longitude",
       y = " ",
       fill = "Probability")

#Get legend
legend <- get_legend(dec_plot)

#Remove legend from December plot
dec_plot <- dec_plot + theme(legend.position = 'none')

#Plotting together
plot_mod <- plot_grid(nov_plot, dec_plot, legend, ncol = 3, nrow = 1,
                            rel_widths = c(1, 1, 0.3))

#Add title
title <- ggdraw()+
  draw_label("Mean crabeater seal distribution\n(ACCESS-OM2-01 - all variables)",
             fontface = "bold", hjust = 0.5)+
  theme(plot.margin = margin(0, 0, 0, 0))

#Putting everything together
final <- plot_grid(title, plot_mod, ncol = 1, 
          rel_heights = c(0.1, 1))

#Saving graph
ggsave(file.path(out_folder, "map_mean_pred_mod.png"), plot = final, 
       device = "png", bg = "white", width = 8.75, height = 7)
```


## Environmental variables from observations 
Finally, we will create a model using all environmental data obtained from observations.  
  
```{r}
#Loading data
obs_env_data <- read_csv(str_subset(file_list, "/obs")) %>% 
  #Setting month as factor and ordered factor
  mutate(month = factor(month))

#List of covariates
covars <- str_subset(names(obs_env_data), "presence|_ocean", negate = T)
```
  
### Building GAM formula
We will run an initial GAM using all available covariates. Sea ice related variables will be fitted individually to each `month` as we expect sea ice to vary in extent between November and December because sea ice should be retreating during this time.  
  
```{r}
# Most complex model
full_model <- presence ~ month + s(year) + s(bottom_slope_deg) + s(dist_shelf_km) + s(dist_coast_km) + s(depth_m) + s(SIC, by = month) + s(SST_degC, by = month) + s(lt_pack_ice, by = month) + s(dist_ice_edge_km, by = month)
```
  
### Splitting data into testing and training
The `prep_data` function in the `useful_functions` script will be used to split our data and to apply all necessary transformations.  
  
```{r}
#Getting training data
obs_data_split <- prep_data(obs_env_data, cat_vars)

#Selecting model data
model_data <- obs_data_split$baked_train %>%
  select(all_of(covars) | "presence")
```
  
### Modelling
Background data will be down-weighted and then we will run the GAM using all covariates available.  
  
```{r}
#Calculating downweights
weights <- down_weights(model_data)

#Full model
full_model_gam <- gam(formula = as.formula(full_model),
    data = model_data,
    family = binomial(link = "cloglog"),
    weights = weights,
    method = "REML", select = T)

#Print summary
summary(full_model_gam)
```  
  
This model outperforms the GAMs trained with environmental data obtained from ACCESS-OM2-01. As we have done before, we will remove any variables that do not contribute much to the model: bottom slope (`bottom_slope_deg`) and distance to the continental shelf (`dist_shelf_km`) and the coast (`dist_coast_km`).  
  
```{r}
# Simplified model
simple_obs_model <- presence ~ month + s(depth_m) + s(SIC, by = month) + s(SST_degC, by = month)+ s(lt_pack_ice, by = month) + s(dist_ice_edge_km, by = month)

#Fitting model
simple_obs_model_gam <- gam(formula = as.formula(simple_obs_model),
    data = model_data,
    family = binomial(link = "cloglog"),
    weights = weights,
    method = "REML", select = T)

#Print summary
summary(simple_obs_model_gam)
```
  
This simplified model explains almost the same variance as the full model.  
  
Although multicollinearity was not identified in the observational dataset, we will assess the effect of keeping either `SST` or `SIC` on the predictive ability of the GAM.  
  
```{r}
# Simplified model no SST
simple_obs_model_noSST <- presence ~ month + s(depth_m) + s(SIC, by = month) + s(lt_pack_ice, by = month) + s(dist_ice_edge_km, by = month)

#Fitting model
simple_obs_model_noSST_gam <- gam(formula = as.formula(simple_obs_model_noSST),
    data = model_data,
    family = binomial(link = "cloglog"),
    weights = weights,
    method = "REML", select = T)

#Print summary
summary(simple_obs_model_noSST_gam)
```
  
Removing `SST` from the model results in a decline in the model predictive ability of about 0.6\%. We will now check results if `SIC` is removed.  
  
```{r}
# Simplified model no SIC
simple_obs_model_noSIC <- presence ~ month + s(depth_m) + s(SST_degC, by = month) + s(lt_pack_ice, by = month) + s(dist_ice_edge_km, by = month)

#Fitting model
simple_obs_model_noSIC_gam <- gam(formula = as.formula(simple_obs_model_noSIC),
    data = model_data,
    family = binomial(link = "cloglog"),
    weights = weights,
    method = "REML", select = T)

#Print summary
summary(simple_obs_model_noSIC_gam)
```
  
The effect of removing `SIC` is much larger on the model predictive ability. Finally, we will apply the same model fitted to the ACCESS data.  
  
We will not apply the same model fitted to the ACCESS data because we did not use `SIC`.  
  
### Comparing models  
We will compare these models using AIC and $r^{2}$. We will summarise everything in the table below.  
  
```{r}
obs_sum_gam <- data.frame(model = c("full_model", "simple_obs_model", "simple_obs_model_noSST",
                                    "simple_obs_model_noSIC"),
                      AIC = c(AIC(full_model_gam), AIC(simple_obs_model_gam), AIC(simple_obs_model_noSST_gam),
                              AIC(simple_obs_model_noSIC_gam)),
                      rsq = c(summary(full_model_gam)$r.sq, summary(simple_obs_model_gam)$r.sq,
                              summary(simple_obs_model_noSST_gam)$r.sq, summary(simple_obs_model_noSIC_gam)$r.sq))

#Rearraging data based on AIC
arrange(obs_sum_gam, AIC)
```
  
The simple model has the highest AIC, and its $r^{2}$, so we will use this to predict crabeater distribution.  
  
### Predictions
We will use the best performing model to predict crabeater seal distribution using mean monthly environmental conditions obtained from observations.  
  
```{r}
#Prediction monthly crabeater seal distribution
mean_pred_obs <- mean_obs_baked %>% 
  mutate(pred = as.vector(predict(simple_obs_model_gam, 
                                  mean_obs_baked, 
                                  type = "response")))

#Converting from data frame to raster
mean_pred_obs_ras <- mean_pred_obs %>%
  #Select relevant variables only
  select(xt_ocean, yt_ocean, pred, month) %>% 
  #Set dimensions
  st_as_stars(dims = c("xt_ocean", "yt_ocean", "month")) %>% 
  #Ensuring month dimension is shown correctly
  st_set_dimensions("month", values = c(11, 12)) %>%
  #Set CRS
  st_set_crs(4326) %>% 
  #Transform to South Pole stereographic
  st_transform(crs = st_crs(3976))

#Saving outputs
#Data frame
mean_pred_obs %>% 
  write_csv(file.path(out_folder, "mean_pred_obs.csv"))
#Saving as R dataset so it can be easily open with readRDS
saveRDS(mean_pred_obs_ras,
        file.path(out_folder, "mean_pred_obs_raster.rds"))
```
  
Now that we have predicted the mean distribution of crabeaters based with our select model, we can plot results for comparison later.  
  
```{r}
#Plotting November distribution
#Prepping data
nov <- mean_pred_obs_ras %>% 
  slice(index = 1, along = "month") 

#Plotting
nov_plot <- ggplot()+
  geom_stars(data = nov)+
  geom_sf(data = antarctica)+
  lims(x = c(0, 4000000))+
  #Set colour palette
  scale_fill_cmocean(name = "haline", direction = -1, 
                     guide = guide_colorbar(barwidth = 1, barheight = 10, 
                                            ticks = FALSE, nbin = 1000, 
                                            frame.colour = "black"), 
                     limits = c(0, 1)) +
  theme_linedraw() +
  theme(panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), 
        legend.position = "none",
        plot.title = element_text(hjust = 0.5)) +
  labs(title = "November",
       x = "Longitude",
       y = "Latitude")

dec <- mean_pred_obs_ras %>% 
  slice(index = 2, along = "month") 

dec_plot <- ggplot() +
  geom_stars(data = dec) +
  geom_sf(data = antarctica)+
  lims(x = c(0, 4000000))+
  scale_fill_cmocean(name = "haline", direction = -1, 
                     guide = guide_colorbar(barwidth = 1, barheight = 10, 
                                            ticks = FALSE, nbin = 1000, 
                                            frame.colour = "black"), 
                     limits = c(0, 1)) +
  theme_linedraw() +
  theme(panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), 
        plot.title = element_text(hjust = 0.5)) +
  labs(title = "December",
       x = "Longitude",
       y = " ",
       fill = "Probability")

#Get legend
legend <- get_legend(dec_plot)

#Remove legend from December plot
dec_plot <- dec_plot + theme(legend.position = 'none')

#Plotting together
plot_obs <- plot_grid(nov_plot, dec_plot, legend, ncol = 3, nrow = 1,
                            rel_widths = c(1, 1, 0.3))

#Add title
title <- ggdraw()+
  draw_label("Mean crabeater seal distribution\n(Observations)",
             fontface = "bold", hjust = 0.5)+
  theme(plot.margin = margin(0, 0, 0, 0))

#Putting everything together
final <- plot_grid(title, plot_obs, ncol = 1, 
                      rel_heights = c(0.1, 1))

#Saving graph
ggsave(file.path(out_folder, "map_mean_pred_obs.png"), plot = final, 
       device = "png", bg = "white", width = 8.75, height = 7)

```

## Differences across sources of environmental data
  
Comparing results from ACCESS-OM2-01 (all available variables) trained GAM with observations trained GAM.  
  
```{r}
diff <- mean_pred_mod_ras-mean_pred_obs_ras

#Plotting November distribution
#Prepping data
nov <- diff %>% 
  slice(index = 1, along = "month") 

#Plotting
nov_plot <- ggplot()+
  geom_stars(data = nov)+
  geom_sf(data = antarctica)+
  lims(x = c(0, 4000000))+
  #Set colour palette
  scale_fill_cmocean(name = "curl",
                     guide = guide_colorbar(barwidth = 1, barheight = 10, 
                                            ticks = FALSE, nbin = 1000, 
                                            frame.colour = "black"), 
                     limits = c(-1, 1)) +
  theme_linedraw() +
  theme(panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), 
        legend.position = "none",
        plot.title = element_text(hjust = 0.5)) +
  labs(title = "November",
       x = "Longitude",
       y = "Latitude")

dec <- diff %>% 
  slice(index = 2, along = "month") 

dec_plot <- ggplot() +
  geom_stars(data = dec) +
  geom_sf(data = antarctica)+
  lims(x = c(0, 4000000))+
  scale_fill_cmocean(name = "curl",
                     guide = guide_colorbar(barwidth = 1, barheight = 10, 
                                            ticks = FALSE, nbin = 1000, 
                                            frame.colour = "black"), 
                     limits = c(-1, 1)) +
  theme_linedraw() +
  theme(panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), 
        plot.title = element_text(hjust = 0.5)) +
  labs(title = "December",
       x = "Longitude",
       y = " ",
       fill = "Probability")

#Get legend
legend <- get_legend(dec_plot)

#Remove legend from December plot
dec_plot <- dec_plot + theme(legend.position = 'none')

#Plotting together
plot_obs_mod <- plot_grid(nov_plot, dec_plot, legend, ncol = 3, nrow = 1,
                            rel_widths = c(1, 1, 0.3))

#Add title
title <- ggdraw()+
  draw_label("Differences in crabeater seal distribution\n(Full ACCESS-OM2-01 vs Observations)",
             fontface = "bold", hjust = 0.5)+
  theme(plot.margin = margin(0, 0, 0, 0))

#Putting everything together
final <- plot_grid(title, plot_obs_mod, ncol = 1, 
                      rel_heights = c(0.1, 1))

#Saving graph
ggsave(file.path(out_folder, "map_comp_pred_obs_mod.png"), plot = final, 
       device = "png", bg = "white", width = 8.75, height = 7)
```
  
Comparing results from ACCESS-OM2-01 (variables match observations) trained GAM with observations trained GAM.  
  
```{r}
diff_2 <- mean_pred_match_obs_ras-mean_pred_obs_ras

#Plotting November distribution
#Prepping data
nov <- diff_2 %>% 
  slice(index = 1, along = "month") 

#Plotting
nov_plot <- ggplot()+
  geom_stars(data = nov)+
  geom_sf(data = antarctica)+
  lims(x = c(0, 4000000))+
  #Set colour palette
  scale_fill_cmocean(name = "curl",
                     guide = guide_colorbar(barwidth = 1, barheight = 10, 
                                            ticks = FALSE, nbin = 1000, 
                                            frame.colour = "black"), 
                     limits = c(-1, 1)) +
  theme_linedraw() +
  theme(panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), 
        legend.position = "none",
        plot.title = element_text(hjust = 0.5)) +
  labs(title = "November",
       x = "Longitude",
       y = "Latitude")

dec <- diff_2 %>% 
  slice(index = 2, along = "month") 

dec_plot <- ggplot() +
  geom_stars(data = dec) +
  geom_sf(data = antarctica)+
  lims(x = c(0, 4000000))+
  scale_fill_cmocean(name = "curl",
                     guide = guide_colorbar(barwidth = 1, barheight = 10, 
                                            ticks = FALSE, nbin = 1000, 
                                            frame.colour = "black"), 
                     limits = c(-1, 1)) +
  theme_linedraw() +
  theme(panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), 
        plot.title = element_text(hjust = 0.5)) +
  labs(title = "December",
       x = "Longitude",
       y = " ",
       fill = "Probability")

#Get legend
legend <- get_legend(dec_plot)

#Remove legend from December plot
dec_plot <- dec_plot + theme(legend.position = 'none')

#Plotting together
plot_obs_mod_lim <- plot_grid(nov_plot, dec_plot, legend, ncol = 3, nrow = 1,
                            rel_widths = c(1, 1, 0.3))

#Add title
title <- ggdraw()+
  draw_label("Differences in crabeater seal distribution\n(Simplified ACCESS-OM2-01 vs Observations)",
             fontface = "bold", hjust = 0.5)+
  theme(plot.margin = margin(0, 0, 0, 0))

#Putting everything together
final <- plot_grid(title, plot_obs_mod_lim, ncol = 1, 
                      rel_heights = c(0.1, 1))

#Saving graph
ggsave(file.path(out_folder, "map_comp_pred_obs_mod_lim.png"), plot = final,
       device = "png", bg = "white", width = 8.75, height = 7)

```
