---
title: "Random Forests"
author: "Denisse Fierro Arcos"
date: "2023-11-15"
output: 
  github_document:
    toc: true
    html_preview: false
---
# Random Forests via SDMtune

Random Forests is one of the most widely used species distribution model algorithm. 

In this project, we will use MaxEnt as one of the models to be considered in our Species Distribution Model ensemble to estimate the distribution of crabeater seals in the recent past.  
  
## Loading libraries

```{r, message=F, warning=F}
library(tidyverse)
library(SDMtune)
library(stars)
library(sf)
library(cmocean)
library(cowplot)
source("useful_functions.R")
```

## Setting up notebook

Selecting an output folder for GAM results exists and getting a list of data files.  
  
```{r}
#Location of folder for outputs
out_folder <- "../../SDM_outputs/RandomForest/Mod_match_obs"
#If folder does not exist, create one
if(!dir.exists(out_folder)){
  dir.create(out_folder, recursive = T)
}

#Get path to files containing data
file_list <- list.files("../../Environmental_Data/", pattern = "Indian", full.names = T)
```
   
### Loading mean environmental conditions from ACCESS-OM2-01

This dataset includes the mean environmental conditions per month (November and December) over the entire period of study (1981 to 2013).

```{r}
mean_model <- read_csv("../../Environmental_Data/ACCESS-OM2-01/All_values_month_ACCESS-OM2-01_env_vars.csv") %>% 
  mutate(month = as.factor(month)) %>% 
  #Drop variables with high multicollinearity
  select(!c(freez_pot_Wm2, bottom_sal_psu, SIT_m))

#List of categorical variables
cat_vars <- "month"

mean_model_baked <- prep_pred(mean_model, cat_vars)
```
    
## Loading layers for plotting
We will extract this layer from the `rnaturalearth` package. We will then reproject this layer to South Polar Stereographic (`EPSG 3976`).  
  
```{r}
#Loading layer
antarctica <- rnaturalearth::ne_countries(continent = "Antarctica",
                                          returnclass = "sf") %>% 
  #Transforming to South Polar Stereographic
  st_transform(3976)
```
  
## Loading environmental data from ACCESS-OM2-01 and setting up variables

We will use the datasets created in the notebook `02_Merging_background_presence_data.Rmd` located within the `Scripts/05_SDMs` folder. These datasets include the crabeater seal observations, background points, and environmental data.  
  
We will also define categorical and continuous explanatory variables.  
  
### Environmental variables matching observations 
First, we will look only at the variables with no multicollinearity. This means that sea surface temperature (`SST`) is excluded even though this variable is available in the observational dataset.  
  
The variable `month` will be included as an ordinal factor in our analysis.  
  
```{r}
#Loading data
mod_match_obs <- read_csv(str_subset(file_list, "match")) %>% 
  #Setting month as factor and ordered factor
  mutate(month = as.factor(month))
```
  
#### Splitting data into testing and training
The `prep_data` function in the `useful_functions` script will be used to split our data and to apply all necessary transformations.
We will then transform the data into SWD ("samples with data") format, which is the required format for inputs used in the `SDMtune` library.   
   
```{r}
#Getting training data
mod_match_obs <- prep_data(mod_match_obs, cat_vars, split = F)

#Applying SWD format to model data
model_data <- mod_match_obs %>% 
  select(!year) %>% 
  sdm_format() %>% 
  trainValTest(test = 0.25, only_presence = T, seed = 42)
```
  
### Modelling
Random Forest has three hyperparameters that can be adjusted to improve model performance:  
1. `mtry`, which refers to the number of variables randomly sampled as candidates at each split   
2. `ntree` is the number of trees that will be grown    
3. `nodesize` is the minimum size of the terminal nodes  
  
Here, we use the `SDMtune` library to test various values combinations for these three hyperparameters. We will identify the "best model" using the `AUC` calculated from the testing dataset.
  
```{r}
#Train model
default_model <- train(method = "RF", data = model_data[[1]])

#Find number of predictors/features used
n_features <- ncol(model_data[[1]]@data)

# Define the hyperparameters to test
hyp_parm <- list(mtry = seq(2, n_features, 2),
                 #number of trees
                 ntree = seq(n_features*10, n_features*100, length.out = 10),
                 #minimum node size
                 nodesize = 1:5)

# Test all the possible combinations with gridSearch
gs_mod_simp <- gridSearch(default_model, hypers = hyp_parm, metric = "auc", 
                          test = model_data[[2]])

default_model
```



