---
title: "Generalised Additive Model"
author: "Denisse Fierro Arcos"
date: "2023-09-22"
output: 
  github_document:
    toc: true
    html_preview: false
---

# Generalised Additive Model (GAM)

GAMs are commonly used in species distribution modelling because of their flexibility. GAMs use smoothing functions to capture non-linear relationships between the dependent variable and the predictors (i.e., independent variables).

In this project, we will use GAMs as one of the models to be considered in our Species Distribution Model ensemble to estimate the distribution of crabeater seals in the recent past.  
  
## Loading libraries

```{r, message=F, warning=F}
library(tidyverse)
library(mgcv)
library(stars)
library(cmocean)
library(cowplot)
source("useful_functions.R")
```


## Setting up notebook

Selecting an output folder for GAM results exists and getting a list of data files.  
  
```{r}
#Location of folder for outputs
out_folder <- "../../SDM_outputs/GAM"
#If folder does not exist, create one
if(!dir.exists(out_folder)){
  dir.create(out_folder, recursive = T)
}

#Get path to files containing data
file_list <- list.files("../../Environmental_Data/", pattern = "Indian", full.names = T)
```
  
### Loading mean environmental conditions from ACCESS-OM2-01

This dataset includes the mean environmental conditions per month (November and December) over the entire period of study (1981 to 2013).

```{r}
mean_model <- read_csv("../../Environmental_Data/ACCESS-OM2-01/All_values_month_ACCESS-OM2-01_env_vars.csv") %>% 
  mutate(month = factor(month),
         month_ordered = factor(month, ordered = T)) %>% 
  #Drop variables with high multicollinearity
  select(!c(freez_pot_Wm2, bottom_sal_psu, SIT_m))

#List of categorical variables
cat_vars <- c("month", "month_ordered")

mean_model_baked <- prep_pred(mean_model, cat_vars)
```
  
## Loading environmental data from ACCESS-OM2-01 and setting up variables

We will use the datasets created in the notebook `02_Merging_background_presence_data.Rmd` located within the `Scripts/05_SDMs` folder. These datasets include the crabeater seal observations, background points, and environmental data.  
  
We will also define categorical and continuous explanatory variables.  
  
### Environmental variables matching observations 
First, we will look only at the variables with no multicollinearity. This means that sea surface temperature (`SST`) is excluded even though this variable is available in the observational dataset.  
  
The variable `month` will be included as an ordinal factor in our analysis.  
  
```{r}
#Loading data
mod_match_obs <- read_csv(str_subset(file_list, "match")) %>% 
  #Setting month as factor and ordered factor
  mutate(month = factor(month),
         month_ordered = factor(month, ordered = T))

#List of covariates
covars <- str_subset(names(mod_match_obs), "presence|_ocean", negate = T)
```
  
#### Building GAM formula
We will run an initial GAM using all available covariates. Sea ice related variables will be fitted individually to each `month` as we expect sea ice to vary in extent between November and December because sea ice should be retreating during this time.  
  
```{r}
# Most complex model
full_model <- presence ~ month + s(year) + s(bottom_slope_deg) + s(dist_shelf_km) + s(dist_coast_km) + s(depth_m) + s(SIC) + s(SIC, by = month_ordered) + s(lt_pack_ice) + s(lt_pack_ice, by = month_ordered) + s(dist_ice_edge_km) + s(dist_ice_edge_km, by = month_ordered)
```
    
#### Splitting data into testing and training
The `prep_data` function in the `useful_functions` script will be used to split our data and to apply all necessary transformations.  
  
```{r}
#Getting training data
mod_match_obs_split <- prep_data(mod_match_obs, cat_vars)

#Selecting model data
model_data <- mod_match_obs_split$baked_train %>%
  select(all_of(covars) | "presence")
```
    
#### Modelling
Background data (`presence` == `0`) will be down-weighted because they do not truly represent absences. The down-weighting applied represents the proportion of presence to background points. In this way, the sum of the weighted background points will be the same as the total of presences.  
  
We will run the GAM using all covariates available and we will downweight background points so they have the same effect as all presences.  

**Other option: Infinitely weighted - 1e3^(1-y) - `model_data %>% mutate(weight = 1e3^(1-presence)) %>% pull(weight)`**
  
```{r}
#Calculating downweights
weights <- down_weights(model_data)

#Full model
full_model_gam <- gam(formula = as.formula(full_model),
    data = model_data,
    family = binomial(link = "cloglog"),
    weights = weights,
    method = "REML", select = T)

#Print summary
summary(full_model_gam)
```
  
Removing variables with effects reduced to near zero (`edf` < 0.001) and that are non-significant: `year`, `bottom_slope_deg` (seafloor slope), `dist_coast_km` (distance to coast in km). After evaluating the diagnostic plots from the model above, the `dist_shelf_km` (distance to continental shelf in km) will also be removed because its effect is negligible.
    
```{r}
#Defining new formula without non-significant covariates
significant_only <-  presence ~ month + s(dist_shelf_km) + s(depth_m) + s(SIC) + s(SIC, by = month_ordered) + s(lt_pack_ice) + s(lt_pack_ice, by = month_ordered) + s(dist_ice_edge_km) + s(dist_ice_edge_km, by = month_ordered)

#Applying new formula
significant_only_gam <- gam(formula = as.formula(significant_only),
    data = model_data,
    family = binomial(link = "cloglog"),
    weights = weights,
    method = "REML", select = T)

summary(significant_only_gam)
```
  
#### Comparing models
The Akaike's Information Criterion (AIC) will be calculated to assess model performance. We will also check the $r^{2}$ values for both models.    

```{r}
#Putting everything in a data frame
sum_gam <- data.frame(model = c("full_model", "significant_only"),
                      AIC = c(AIC(full_model_gam), AIC(significant_only_gam)),
                      rsq = c(summary(full_model_gam)$r.sq, summary(significant_only_gam)$r.sq))
#Checking results
sum_gam
```
  
These models performed almost the same, so we will keep the most parsimonious model (i.e., `significant_only`), which has a slightly lower AIC and slightly higher $r^{2}$.  We will include `SST` to this model and check the effect of `SST` on model performance. The smoothing for this variable will vary by month as we have done with sea ice related variables.  
  
```{r}
significant_only_SST <-  presence ~ month + s(dist_shelf_km) + s(depth_m) + s(SIC) + s(SIC, by = month_ordered) + s(lt_pack_ice) + s(lt_pack_ice, by = month_ordered) + s(dist_ice_edge_km) + s(dist_ice_edge_km, by = month_ordered) + s(SST_degC) + s(SST_degC, by = month_ordered)

#Applying new formula
significant_only_SST_gam <- gam(formula = as.formula(significant_only_SST),
    data = model_data,
    family = binomial(link = "cloglog"),
    weights = weights,
    method = "REML", select = T)

#Check results
summary(significant_only_SST_gam)
```
  
The model that includes `SST` explains almost 9\% of variability in the data, which is higher than the previous models. We can see that now sea ice concentration (`SIC`) has become insignificant, and this is likely because of its high correlation to `SST`. We will keep the `SIC` by month because it is highly significant. We will remove this variable, together with distance to the shelf (`dist_shelf_km`) because its effect is near zero (`edf` < 0.001) and it is also non-significant.  
  
```{r}
significant_only_no_SIC <-  presence ~ month + s(depth_m) + s(SIC, by = month_ordered) + s(lt_pack_ice) + s(lt_pack_ice, by = month_ordered) + s(dist_ice_edge_km) + s(dist_ice_edge_km, by = month_ordered) + s(SST_degC) + s(SST_degC, by = month_ordered)

#Applying new formula
significant_only_no_SIC_gam <- gam(formula = as.formula(significant_only_no_SIC),
    data = model_data,
    family = binomial(link = "cloglog"),
    weights = weights,
    method = "REML", select = T)

#Check results
summary(significant_only_no_SIC_gam)
```
  
The performance of this reduced model is basically the same as the previous one. We will calculate AIC and extract $r^{2}$ values and add them to our summary table.  
  
```{r}
#Adding results to summary table
sum_gam <- data.frame(model = c("significant_only_SST", "significant_only_no_SIC"),
                      AIC = c(AIC(significant_only_SST_gam), AIC(significant_only_no_SIC_gam)),
                      rsq = c(summary(significant_only_SST_gam)$r.sq, summary(significant_only_no_SIC_gam)$r.sq)) %>% 
  bind_rows(sum_gam, .) %>% 
  #Arrange in increasing order by AIC
  arrange(AIC)

#Checking results
sum_gam
```
  
We can see that the best performing model based on AIC is the last one we tested, which includes `SST`, excludes `SIC`, but keeps `SIC` by month (`significant_only_SST` in table above). We can check if the performance of this model differs significantly from the model that did not include `SST` (`significant_only` in table above). To do this, we will performance an ANOVA test.  
  
```{r}
anova(significant_only_gam, significant_only_no_SIC_gam, test = "Chisq")
```
  
We can see these models differ significantly, so we will use the `significant_only_no_SIC` to predict crabeater distribution.  
  
#### Predictions
We will use the best performing model to predict crabeater seal distribution using mean monthly environmental conditions obtained from ACCESS-OM2-01.  
  
```{r}
#Prediction monthly crabeater seal distribution
mean_pred_match_obs <- mean_model_baked %>% 
  mutate(pred = as.vector(predict(significant_only_no_SIC_gam, 
                                  mean_model_baked, 
                                  type = "response")))

#Converting from data frame to raster
mean_pred_match_obs_ras <- mean_pred_match_obs %>%
  #Select relevant variables only
  select(xt_ocean, yt_ocean, pred, month) %>% 
  #Set dimensions
  st_as_stars(dims = c("xt_ocean", "yt_ocean", "month")) %>% 
  #Ensuring month dimension is shown correctly
  st_set_dimensions("month", values = c(11, 12)) %>%
  #Set CRS
  st_set_crs(4326) %>% 
  #Transform to South Pole stereographic
  st_transform(crs = st_crs(3976))


#Saving outputs
#Data frame
mean_pred_match_obs %>% 
  write_csv(file.path(out_folder, "mean_pred_match_obs.csv"))
#NetCDF file
write_stars(mean_pred_match_obs_ras, 
            file.path(out_folder, "mean_pred_match_obs_raster.nc"),
            driver = "netcdf")
```


```{r}
#Plotting november distribution
nov <- mean_pred_match_obs_ras %>% 
  slice(index = 1, along = "month") 

nov_plot <- ggplot()+
  geom_stars(data = nov)+
  scale_fill_cmocean(name = "ice", direction = -1, 
                     guide = guide_colorbar(barwidth = 1, barheight = 10, 
                                            ticks = FALSE, nbin = 1000, 
                                            frame.colour = "black"), 
                     limits = c(0, 1)) +
  theme_linedraw() +
  theme(panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  labs(title = "November",
       x = "Longitude",
       y = "Latitude",
       fill = "Probability")

dec <- mean_pred_match_obs_ras %>% 
  slice(index = 2, along = "month") 

dec_plot <- ggplot() +
  geom_stars(data = dec) +
  scale_fill_cmocean(name = "ice", direction = -1, 
                     guide = guide_colorbar(barwidth = 1, barheight = 10, 
                                            ticks = FALSE, nbin = 1000, 
                                            frame.colour = "black"), 
                     limits = c(0, 1)) +
  theme_linedraw() +
  theme(panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  labs(title = "December",
       x = "Longitude",
       y = "Latitude",
       fill = "Probability")

#Plotting together
nov_plot+dec_plot

```



  
### All environmental variables available in ACCESS-OM2-01
We will now use the full range of covariates available in the ACCESS-OM2-01 model. We will test if the inclusion of additional environmental variables improve predictive ability. We will follow the same approach as we did before, first we will test a model with all covariates and exclude any with non-significant effects.  
  
We tested this dataset for multicollinearity, so we will get the names of variables with low VIF plus SST as this was found to be an important covariate before.  
  
```{r}
#Loading data
full_mod <- read_csv(str_subset(file_list, "model_env_pres")) %>% 
  #Setting month as factor and ordered factor
  mutate(month = factor(month),
         month_ordered = factor(month, ordered = T))

#List of variables to be extracted from main dataset
mod_vars <- read_csv("../../Environmental_Data/ACCESS-OM2-01/Obs_BG_5x_Indian_weaning_LowVIF.csv") %>% 
  select(!c(sector, zone, season_year:decade)) %>% 
  names() %>% 
  append(c("SST_degC", "month_ordered"))

#Extracting relevant columns only
full_mod <- full_mod %>% 
  select(all_of(mod_vars))

#List of environmental covariates to be used in model
covars <- str_subset(names(full_mod), "presence|_ocean", negate = T)
```
  
#### Building GAM formula
We will run an initial GAM using all available covariates. Non-static variables (i.e., variables that vary with time) will be fitted individually to each `month` as we expect them to vary between November and December as we move from spring to summer.  
  
```{r}
# Most complex model
full_model <- presence ~ month + s(year) + s(bottom_slope_deg) + s(dist_shelf_km) + s(dist_coast_km) + s(depth_m) + s(SIC) + s(SIC, by = month_ordered) + s(bottom_temp_degC) + s(bottom_temp_degC, by = month_ordered) + s(SSS_psu) + s(SSS_psu, by = month_ordered) + s(vel_lat_surf_msec) + s(vel_lat_surf_msec, by = month_ordered) + s(vel_lat_bottom_msec) + s(vel_lat_bottom_msec, by = month_ordered) + s(vel_lon_surf_msec) + s(vel_lon_surf_msec, by = month_ordered) + s(vel_lon_bottom_msec) + s(vel_lon_bottom_msec, by = month_ordered) + s(lt_pack_ice) + s(lt_pack_ice, by = month_ordered) + s(dist_ice_edge_km) + s(dist_ice_edge_km, by = month_ordered)
```
    
#### Splitting data into testing and training
The `prep_data` function in the `useful_functions` script will be used to split our data and to apply all necessary transformations.  
  
```{r}
#Getting training data
full_mod_split <- prep_data(full_mod, cat_vars)

#Selecting model data
model_data <- full_mod_split$baked_train %>%
  select(all_of(covars) | "presence")
```
  
#### Modelling
Background data will be down-weighted and then we will run the GAM using all covariates available.  
  
```{r}
#Calculating downweights
weights <- down_weights(model_data)

#Full model
full_model_gam <- gam(formula = as.formula(full_model),
    data = model_data,
    family = binomial(link = "cloglog"),
    weights = weights,
    method = "REML", select = T)

#Print summary
summary(full_model_gam)
```  
  
This model explains more of the variability in our data than the model that uses variables for which we also have observations. We will remove the variables with a low contribution (`edf` $\sim 0.001$). This includes `year`, bottom slope (`bottom_slope_deg`), zonal water velocity at the surface (`vel_lat_surf_msec`, but we will keep monthly fits), meridional water velocity at the bottom (`vel_lon_bottom_msec`), and the monthly fits for meridional water velocity at the surface(`vel_lon_surf_msec`), bottom zonal velocity (`vel_lat_bottom_msec`) and sea ice concentration (`SIC`).

We will defined this simplified model and test it below.  
  
```{r}
# Simplified model
simpler_model <- presence ~ s(dist_shelf_km) + s(dist_coast_km) + s(depth_m) + s(SIC) + s(bottom_temp_degC) + s(bottom_temp_degC, by = month_ordered) + s(SSS_psu) + s(SSS_psu, by = month_ordered) + s(vel_lat_surf_msec, by = month_ordered) + s(vel_lat_bottom_msec) + s(vel_lon_surf_msec) + s(vel_lon_bottom_msec, by = month_ordered) + s(lt_pack_ice) + s(lt_pack_ice, by = month_ordered) + s(dist_ice_edge_km) + s(dist_ice_edge_km, by = month_ordered)

#Applying new simplified formula
simpler_model_gam <- gam(formula = as.formula(simpler_model),
    data = model_data,
    family = binomial(link = "cloglog"),
    weights = weights,
    method = "REML", select = T)

#Print summary
summary(simpler_model_gam)
```
  
We can see that this simplified model performs slightly better than the original model including all environmental variables available.  
  
We will now explore the influence of `SST` as it was identified as an influential variable in the previous dataset.  
  
```{r}
# Simplified model with SST
simpler_model_SST <- presence ~ month + s(dist_shelf_km) + s(dist_coast_km) + s(depth_m) + s(SIC) + s(bottom_temp_degC) + s(bottom_temp_degC, by = month_ordered) + s(SSS_psu) + s(SSS_psu, by = month_ordered) + s(vel_lat_surf_msec, by = month_ordered) + s(vel_lat_bottom_msec) + s(vel_lon_surf_msec) + s(vel_lon_bottom_msec, by = month_ordered) + s(lt_pack_ice) + s(lt_pack_ice, by = month_ordered) + s(dist_ice_edge_km) + s(dist_ice_edge_km, by = month_ordered) + s(SST_degC) + s(SST_degC, by = month_ordered)

#Applying new simplified formula
simpler_model_SST_gam <- gam(formula = as.formula(simpler_model_SST),
    data = model_data,
    family = binomial(link = "cloglog"),
    weights = weights,
    method = "REML", select = T)

#Print summary
summary(simpler_model_SST_gam)
```
  
This model is slightly worse than when `SST` was not included. This is likely because `SST` was highly correlated to `SIC`. Therefore, we will not use `SST` to predict crabeater seal distribution.  
  
#### Comparing models
Once again, we will calculate AIC for each model and obtain the coefficient of determination ($r^{2}$) to assess model performance.   

```{r}
full_sum_gam <- data.frame(model = c("full_model", "simpler_model", "simpler_model_SST"),
                      AIC = c(AIC(full_model_gam), AIC(simpler_model_gam), AIC(simpler_model_SST_gam)),
                      rsq = c(summary(full_model_gam)$r.sq, summary(simpler_model_gam)$r.sq, summary(simpler_model_SST_gam)$r.sq))

#Arranging models from lowest to highest AIC
full_sum_gam %>% 
  arrange(AIC)
```
  
Based on AIC and $r^{2}$, we can see that the simplified model is the best performing of the three models tested here.  
  
### Environmental variables from observations 
Finally, we will create a model using all environmental data obtained from observations.  
  
```{r}
#Loading data
obs_env_data <- read_csv(str_subset(file_list, "/obs")) %>% 
  #Setting month as factor and ordered factor
  mutate(month = factor(month),
         month_ordered = factor(month, ordered = T))

#List of covariates
covars <- str_subset(names(obs_env_data), "presence|_ocean", negate = T)
```
  
#### Building GAM formula
We will run an initial GAM using all available covariates. Sea ice related variables will be fitted individually to each `month` as we expect sea ice to vary in extent between November and December because sea ice should be retreating during this time.  
  
```{r}
# Most complex model
full_model <- presence ~ month + s(year) + s(bottom_slope_deg) + s(dist_shelf_km) + s(dist_coast_km) + s(depth_m) + s(SIC) + s(SIC, by = month_ordered) + s(SST_degC) + s(SST_degC, by = month_ordered) + s(lt_pack_ice) + s(lt_pack_ice, by = month_ordered) + s(dist_ice_edge_km) + s(dist_ice_edge_km, by = month_ordered)
```
  
#### Splitting data into testing and training
The `prep_data` function in the `useful_functions` script will be used to split our data and to apply all necessary transformations.  
  
```{r}
#Getting training data
obs_data_split <- prep_data(obs_env_data, cat_vars)

#Selecting model data
model_data <- obs_data_split$baked_train %>%
  select(all_of(covars) | "presence")
```
  
#### Modelling
Background data will be down-weighted and then we will run the GAM using all covariates available.  
  
```{r}
#Calculating downweights
weights <- down_weights(model_data)

#Full model
full_model_gam <- gam(formula = as.formula(full_model),
    data = model_data,
    family = binomial(link = "cloglog"),
    weights = weights,
    method = "REML", select = T)

#Print summary
summary(full_model_gam)
```  
  
This model outperforms the GAMs trained with environmental data obtained from ACCESS-OM2-01. As we have done before, we will remove any variables that do not contribute much to the model: `year`, bottom slope (`bottom_slope_deg`) and long-term pack ice stability (`lt_pack_ice`), which is non-significant. We will run the model once more to check if we can improve it by simplifying it.  
  
```{r}
# Simplified model
simple_obs_model <- presence ~ month + s(dist_shelf_km) + s(dist_coast_km) + s(depth_m) + s(SIC) + s(SIC, by = month_ordered) + s(SST_degC) + s(SST_degC, by = month_ordered) + s(dist_ice_edge_km) + s(dist_ice_edge_km, by = month_ordered)

#Fitting model
simple_obs_model_gam <- gam(formula = as.formula(simple_obs_model),
    data = model_data,
    family = binomial(link = "cloglog"),
    weights = weights,
    method = "REML", select = T)

#Print summary
summary(simple_obs_model_gam)
```
  
#### Comparing models  
We will compare these models using AIC and $r^{2}$. We will summarise everything in the table below.  
  
```{r}
obs_sum_gam <- data.frame(model = c("full_model", "simple_obs_model"),
                      AIC = c(AIC(full_model_gam), AIC(simple_obs_model_gam)),
                      rsq = c(summary(full_model_gam)$r.sq, summary(simple_obs_model_gam)$r.sq))

#Rearraging data based on AIC
arrange(obs_sum_gam, AIC)
```
  
The simpler model has a lower AIC, and its $r^{2}$ value is quite similar to that from the more complex model. Since we are aiming to keep the most parsimonious model, we will keep the simplified version to predict crabeater seal distributions.  








```{r}
test_pred <- c(predict.gam(mod_gam2, newdata = mod_match_obs_split$baked_test, type = "response")) 

test_pred <- mod_match_obs_split$baked_test %>% 
  select(xt_ocean, yt_ocean, presence) %>% 
  mutate(pred = test_pred)
```





```{r}
test_pred %>% 
  filter(presence == 1) %>% 
  ggplot(aes(xt_ocean, yt_ocean))+
  geom_point()
```

```{r}
test_pred %>% 
  filter(pred >= 0.5) %>% 
  ggplot(aes(xt_ocean, yt_ocean))+
  geom_point(aes(color = pred))+
  scale_color_viridis_c(direction =  -1)
```


```{r}
test_pred %>% 
  filter(pred < 0.5) %>% 
  ggplot(aes(xt_ocean, yt_ocean))+
  geom_point(aes(color = pred))+
  scale_color_viridis_c(direction =  -1)
```




```{r}
#Selecting variables available as 
covars_match_obs <- str_subset(names(mod_match_obs), "presence|_ocean", negate = T)
```



```{r}
gam_mod <- gen_additive_mod(mode = "regression", select_features = T,
                            adjust_deg_free = 5) %>%
  set_engine("mgcv", family = binomial(link = "cloglog"), method = "REML")

gam_fit <- gam_mod %>% 
  fit(as.formula(gam_forms[1]), data = model_data)

tidy(gam_fit)
```


```{r}
gam_mod <- gen_additive_mod(mode = "regression", select_features = T) %>%#,
                            # adjust_deg_free = 1) %>% 
  set_engine("mgcv", family = binomial(link = "cloglog"), method = "REML")

gam_fit <- gam_mod %>% 
  fit(as.formula(gam_forms[1]), data = model_data)

tidy(gam_fit)
```


  
  
```{r}
library(parsnip)
library(tune)
library(dials)

#Creating recipe - steps to be followed when pre-processing data
recipe_bg <- recipe(presence ~ ., mod_match_obs_split$train) %>%
  #Coordinates (xt_ocean and yt_ocean) are not pre-processed
  add_role(xt_ocean, yt_ocean, new_role = "no_prepro") %>% 
  add_role(cat_vars, new_role = "no_prepro") %>% 
  #Scaled and center all predictors - exclude coordinates
  step_center(all_predictors(), -has_role("no_prepro")) %>% 
  step_scale(all_predictors(), -has_role("no_prepro")) %>% 
  prep()

prepro <- recipe_bg %>% 
  bake(mod_match_obs_split$train)

test_prep <- recipe_bg %>% 
  bake(mod_match_obs_split$test)

folds <- vfold_cv(prepro, v = 5)

gam_params <- parameters(adjust_deg_free())

# cv <- trainControl(method = "repeatedcv", number = 10, repeats = 5)
# hyper_grid <- expand.grid(df = 2:5)#formula = gam_forms$gam_forms)
gam_grid <- grid_max_entropy(gam_params, size = 5)

gam <- gen_additive_mod(mode = "regression", 
                        adjust_deg_free = tune()) %>% 
  set_engine("mgcv") 

gam_wf <- workflow() %>% 
  add_variables(outcomes = presence, predictors = env_non_cat) %>% 
  add_model(gam, formula = as.formula(gam_forms[1,])) #%>% 
  # add_recipe(formula = as.formula(gam_forms[1,]))
  # fit(as.formula(gam_forms[1,]), data = mod_match_obs_split$train)

spline_gam <- tune_grid(model = gam, gam_wf,
                        resamples = folds, 
                        grid = gam_grid, metrics = metric_set(mae, rmse, rsq))

show_best(spline_gam, metric = "rmse") 

params <- select_best(spline_gam, metric = "rmse") %>% 
  select(adjust_deg_free)#, maximize = F)

gam_end <- gam %>% 
  finalize_model(parameters = params)

gam_end %>% 
  fit(formula = as.formula(gam_forms[1,]),
      data = prepro) %>% 
  predict(new_data = test_prep) %>% 
  bind_cols(mod_match_obs_split$test) %>% View()

# gam_fit <- train(recipe_bg, data = mod_match_obs_split$train, 
#                  method = "gam", trControl = cv, tuneGrid = hyper_grid,
#                  metric = "RMSE")
```

  
## Loading environmental data
  
```{r}
#Env data to predict
evaluation <- read_csv("../../Environmental_Data/ACCESS-OM2-01/All_values_ACCESS-OM2-01_env_vars.csv") %>% 
  select(any_of(names(mod_match_obs_split$baked_train)))
```





```{r}
library(parsnip)
library(workflows)
library(tidymodels)

crab_folds <- vfold_cv(model_data, v = 5)

spline_rec <- recipe(presence ~ ., data = model_data) %>% 
  step_ns(covars, deg_free = tune("df"))

gam_pars <- gen_additive_mod(adjust_deg_free = tune()) %>% 
  set_engine("mgcv") %>% 
  set_mode("regression") 

spline_grid <- expand.grid(df = 1:5)

gam_simple_wf <- workflow() %>% 
  add_variables(outcomes = presence, predictors = covars) %>% 
  add_model(gam_pars, formula = as.formula(gam_formula))

spline_res <- tune_grid(gam_simple_wf, crab_folds, grid = spline_grid)

%>% 
  fit(data = model_data) %>% 
  add_case_weights(weights) #%>%
  # extract_fit_engine()



gam_simple_wf %>% 
  workflow_map(resamples = crab_folds, grid = 10,
               metrics = metric_set(rmse), verbose = T)


gam_pred <- gam_simple_wf %>%
  predict(mod_match_obs_split$baked_test %>% 
            select(all_of(covars) | "presence"), type = "response") %>%
  as.vector() %>% 
  bind_cols(mod_match_obs_split$baked_test) %>% 
  rename("pred"="...1")

gam_pred %>% 
  roc_auc(presence, pred)#, case_weights = weights)


gam_sel <- gen_additive_mod(select_features = tune(), adjust_deg_free = tune()) %>% 
  set_engine("mgcv") %>% 
  set_mode("regression")

gam_sel_wf <- workflow() %>% 
  add_variables(outcomes = presence, predictors = covars) %>% 
  add_model(gam_pars, formula = as.formula(gam_formula))



gam_sel %>% 
  workflow
  tune_grid(resamples = crab_folds)

%>% 
  fit(data = model_data) %>% 
  add_case_weights(weights) %>% 
  extract_fit_engine()


gam_res <- 


show_best(gam_res, metric = "rmse")
```


