---
title: "Exploratory Data Analysis - Environmental variable selection"
author: "Denisse Fierro Arcos"
date: "2023-07-13"
output:
  github_document:
    toc: true
    html_preview: false
---

# Environmental variable selection

Before developing a Generalised Additive Model (GAM) to estimate the distribution of crabeater seals in East Antarctica, we will check all that the 17 environmental variables identified as potential predictors of crabeater seal distribution are not highly correlated with each other. This is known as multicollinearity and it can cause problems with regression analyses as highly correlated variables do not provide independent information to the regression model and this can cause problems when fitting the model.  
  
To identify multicollinearity, we will calculate the correlation among all environmental variables. High correlation (r > 0.75) can be an indicator of multicollinearity. We will follow this by applying a linear regression using all environmental variables as predictors and the `presence` column as the response (a value of `1` in this columns indicates an observation while `0` refers to background data). We will use these results to calculte the variance inflation factor (VIF). The VIF is the most common way to detect multicollinearity with values above 5 indicating a strong correlation across variables.  
  
Finally, to help us decide which environmental variables we will keep if multicollinearity is detected, we will use the results from the linear regression. We will also plot the density distribution of the environmental variables between observation and background points to help identify which variable to keep as a predictor. Variables that appear to show a larger difference in their density distribution will be preferred as predictors as they capture differences in environmental conditions detected at seal sighting location in comparison to background points.  

## Loading libraries

```{r libraries, results = "hide", message = F, warning = F}
library(tidyverse)
library(GGally)
library(car)
library(cowplot)
```


## Loading observations and background data

We will load these two datasets and merge them together before running our exploratory analysis.

```{r data}
crab_ind <- read_csv("../../Environmental_Data/ACCESS-OM2-01/unique_crabeater_obs_all_env.csv") %>% 
  #Selecting observations for the Indian sector during the weaning period
  filter(str_detect(sector, "Indian") & life_stage == "weaning") %>% 
  #Binding to background data
  bind_rows(read_csv("../../Environmental_Data/ACCESS-OM2-01/unique_background_20x_obs_all_env.csv")) %>% 
  select(!c(event_date:source))

#Checking data structure
glimpse(crab_ind)
```

## Extracting environmental variables
We will create a new variable to facilitate this exploratory analysis.

```{r}
#Checking correlation between numeric variables
num_var <- crab_ind %>% 
  select(bottom_slope_deg:krill_ggp) %>% 
  #Removing rows containing NA values
  drop_na()
```


### Calculating and plotting correlation matrix

We will calculate the Spearman correlations across all variables. We use Spearman because it does not assume relationships across variables are linear.

```{r corrplot}
#Calculating correlations and plotting results
ggcorr(num_var, method = c("pairwise", "spearman"), 
       label = T, label_size = 3, angle = -15)+
  theme(legend.justification = c(1, 0),
        legend.position = c(0.5, 0.7),
        legend.direction = "horizontal") +
  guides(fill = guide_colorbar(barwidth = 9, 
                               barheight = 1, 
                               title.position = "top", 
                               title.hjust = 0.5, 
                               title = "Spearman correlation"))
```
There are a number of variables with a strong correlation (0.75 or above). This suggests that multicollinearity would be an issue in our model if we were to include all variables as predictors. We will continue the analysis by calculating the variance inflation factor (VIF). This will help us identify the variables that we should consider excluding from the model due to their lack of independence with one or more predictor variables.


## Calculating linear regression between predicting and response variable
The response variable will be the `presence` column, which identifies observations and background data. The predicting variables are all the variables shown in the correlation plot above.

```{r lm_prep}
# First we will define the formula to be used in the linear regression. We will use the names of all variables in the num_var data frame
form_lm <- paste("presence", paste(colnames(num_var), sep = "+", 
                                   collapse = "+"), 
                 sep = " ~ ")

#Fitting linear model
model <- lm(as.formula(form_lm), data = crab_ind)

#Checking results
summary(model)
```

We can see that not all variables under consideration were found to have a significant correlation to the `presence` column. We can now calculate VIF values for all predicting variables.


## Calculating and plotting VIF values

```{r vif}
#Calculating VIF values for all predictors in the linear regression
vif_vals <- vif(model)

#Plotting results as a horizontal bar plot
par(mar = c(2, 10, 2, 2))
barplot(vif_vals, horiz = T, las = 1)
#We will add a vertical line at 5 to identify variables with large VIF values
abline(v = 5)
```

Freezing potential (`freez_pot_Wm2`) and sea surface temperature (`SST_degC`) are the two factors with the largest VIF values. Neither of these factors are significant in the model. We will check the density distribution of these variables between observations and backgrounds to help us decide which variable to remove from further analyses.

```{r}
#Density distribution for freezing potential
den1 <- crab_ind %>% 
  drop_na() %>% 
  ggplot(aes(x = freez_pot_Wm2, fill = factor(presence)))+
  geom_density(alpha = 0.5)+
  scale_fill_brewer(palette = "Dark2")+
  theme_bw()

#Density distribution for SST
den2 <- crab_ind %>% 
  drop_na() %>% 
  ggplot(aes(x = SST_degC, fill = factor(presence)))+
  geom_density(alpha = 0.5)+
  scale_fill_brewer(palette = "Dark2")+
  theme_bw()

#Plotting together
plot_grid(den1, den2, nrow = 2)
```

The density distribution of these two variables are almost a mirror image of each other. Neither were considered to be statistically significant in the linear regression. Looking at the correlation plot, we can see that freezing potential has strong correlation with a larger number of variables that SST, so we will drop freezing potential and run the analysis again.


## Correlation matrix without freezing potential

```{r}
#Removing freezing potential from num_var data frame and plot correlations again
num_var <- num_var %>% 
  #Removing rows containing NA values
  drop_na() %>% 
  select(!freez_pot_Wm2)

#Calculating correlations and plotting results
ggcorr(num_var, method = c("pairwise", "spearman"), 
       label = T, label_size = 3, angle = -15)+
  theme(legend.justification = c(1, 0),
        legend.position = c(0.5, 0.7),
        legend.direction = "horizontal") +
  guides(fill = guide_colorbar(barwidth = 9, 
                               barheight = 1, 
                               title.position = "top", 
                               title.hjust = 0.5, 
                               title = "Spearman correlation"))
```

## Linear regression without freezing potential 

High correlations are still detected, so we will run the linear regression and calculate VIF once again.

```{r}
#Linear regression formula
form_lm <- paste("presence", 
                 paste(colnames(num_var), sep = "+", collapse = "+"), 
                 sep = " ~ ")

#Fitting linear model
model <- lm(as.formula(form_lm), data = crab_ind)

#Checking results
summary(model)
```

## Calculating VIF without freezing potential

```{r}
#Calculating VIF values for all predictors in the linear regression
vif_vals <- vif(model)

#Plotting results as a horizontal bar plot
par(mar = c(2, 10, 2, 2))
barplot(vif_vals, horiz = T, las = 1)
#We will add a vertical line at 5 to identify variables with large VIF values
abline(v = 5)
```

By removing the freezing potential, the VIF value for SST decrease in about 90\%. We will now consider the bottom salinity (`bottom_sal_PSU`) and the sea ice concentration (`SIC`). In this case, both variables were found to be significant in the linear regression. We will now look at the density distribution of these variables.

```{r}
#Density distribution for freezing potential
den1 <- crab_ind %>% 
  drop_na() %>% 
  ggplot(aes(x = bottom_sal_psu, fill = factor(presence)))+
  geom_density(alpha = 0.5)+
  scale_fill_brewer(palette = "Dark2")+
  theme_bw()

#Density distribution for SST
den2 <- crab_ind %>% 
  drop_na() %>% 
  ggplot(aes(x = SIC, fill = factor(presence)))+
  geom_density(alpha = 0.5)+
  scale_fill_brewer(palette = "Dark2")+
  theme_bw()

#Plotting together
plot_grid(den1, den2, nrow = 2)
```

There is a larger difference in the density distribution of SIC, while the bottom salinity shows a very similar distribution between background and observations. Since there is not a large difference in the bottom salinity, this may not provide useful information to our model, so we will remove it and run the analysis once more.


## Correlation matrix without bottom salinity

```{r}
#Removing freezing potential from num_var data frame and plot correlations again
num_var <- num_var %>% 
  #Removing rows containing NA values
  drop_na() %>% 
  select(!bottom_sal_psu)

#Calculating correlations and plotting results
ggcorr(num_var, method = c("pairwise", "spearman"), 
       label = T, label_size = 3, angle = -15)+
  theme(legend.justification = c(1, 0),
        legend.position = c(0.5, 0.7),
        legend.direction = "horizontal") +
  guides(fill = guide_colorbar(barwidth = 9, 
                               barheight = 1, 
                               title.position = "top", 
                               title.hjust = 0.5, 
                               title = "Spearman correlation"))
```

## Linear regression without bottom salinity 

High correlations are still present. SIC appears to still be an issue, so we will run the linear regression and calculate VIF once again.

```{r}
#Linear regression formula
form_lm <- paste("presence", 
                 paste(colnames(num_var), sep = "+", collapse = "+"), 
                 sep = " ~ ")

#Fitting linear model
model <- lm(as.formula(form_lm), data = crab_ind)

#Checking results
summary(model)
```
  
## Calculating VIF without bottom salinity

```{r}
#Calculating VIF values for all predictors in the linear regression
vif_vals <- vif(model)

#Plotting results as a horizontal bar plot
par(mar = c(2, 10, 2, 2))
barplot(vif_vals, horiz = T, las = 1)
#We will add a vertical line at 5 to identify variables with large VIF values
abline(v = 5)
```
  
SIC and SST still show a high level of multicollinearity. Both variables are significant in the linear regression, but given that there are larger differences in the density distribution between background points and observations for SIC (see figures in steps above) than for SST, we will drop SST. We will remove SST and run the analysis once more.  
  
## Correlation matrix without SST

```{r}
#Removing freezing potential from num_var data frame and plot correlations again
num_var <- num_var %>% 
  #Removing rows containing NA values
  drop_na() %>% 
  select(!SST_degC)

#Calculating correlations and plotting results
ggcorr(num_var, method = c("pairwise", "spearman"), 
       label = T, label_size = 3, angle = -15)+
  theme(legend.justification = c(1, 0),
        legend.position = c(0.5, 0.7),
        legend.direction = "horizontal") +
  guides(fill = guide_colorbar(barwidth = 9, 
                               barheight = 1, 
                               title.position = "top", 
                               title.hjust = 0.5, 
                               title = "Spearman correlation"))
```
  
Multicollinearity may still be present as correlation values of |0.9| can still be seen in the figure above.  
  
## Linear regression without SST

```{r}
#Linear regression formula
form_lm <- paste("presence",
                 paste(colnames(num_var), sep = "+", collapse = "+"),
                 sep = " ~ ")

#Fitting linear model
model <- lm(as.formula(form_lm), data = crab_ind)

#Checking results
summary(model)
```

## Calculating VIF without SST

```{r}
#Calculating VIF values for all predictors in the linear regression
vif_vals <- vif(model)

#Plotting results as a horizontal bar plot
par(mar = c(2, 10, 2, 2))
barplot(vif_vals, horiz = T, las = 1)
#We will add a vertical line at 5 to identify variables with large VIF values
abline(v = 5)
```
  
Multicollinearity is still present between SIC, distance to continental shelf (`dist_shelf_km`) and krill habitat (`krill_ggp`). We will plot their density distribution.  
  
```{r}
#Density distribution for sea ice thickness
den1 <- crab_ind %>% 
  drop_na() %>% 
  ggplot(aes(x = krill_ggp, fill = factor(presence)))+
  geom_density(alpha = 0.5)+
  scale_fill_brewer(palette = "Dark2")+
  theme_bw()

#Density distribution for SIC
den2 <- crab_ind %>% 
  drop_na() %>% 
  ggplot(aes(x = SIC, fill = factor(presence)))+
  geom_density(alpha = 0.5)+
  scale_fill_brewer(palette = "Dark2")+
  theme_bw()

#Plotting together
plot_grid(den1, den2, nrow = 2)
```
  
Since SIC was non-significant in the linear regression when krill was also considered, we will remove SIC and test for multicollinearity again.  
  
## Correlation matrix without SIC

```{r}
#Removing freezing potential from num_var data frame and plot correlations again
num_var <- num_var %>% 
  #Removing rows containing NA values
  drop_na() %>% 
  select(!SIC)

#Calculating correlations and plotting results
ggcorr(num_var, method = c("pairwise", "spearman"), 
       label = T, label_size = 3, angle = -15)+
  theme(legend.justification = c(1, 0),
        legend.position = c(0.5, 0.7),
        legend.direction = "horizontal") +
  guides(fill = guide_colorbar(barwidth = 9, 
                               barheight = 1, 
                               title.position = "top", 
                               title.hjust = 0.5, 
                               title = "Spearman correlation"))
```
  
There are still some variables that are highly correlated. We will calculate VIF on the remaining variables.  
  
## Linear regression without SIC

```{r}
#Linear regression formula
form_lm <- paste("presence", 
                 paste(colnames(num_var), sep = "+", collapse = "+"), 
                 sep = " ~ ")

#Fitting linear model
model <- lm(as.formula(form_lm), data = crab_ind)

#Checking results
summary(model)
```

## Calculating VIF without SIC

```{r}
#Calculating VIF values for all predictors in the linear regression
vif_vals <- vif(model)

#Plotting results as a horizontal bar plot
par(mar = c(2, 10, 2, 2))
barplot(vif_vals, horiz = T, las = 1)
#We will add a vertical line at 5 to identify variables with large VIF values
abline(v = 5)
```
  
Distance to continental shelf (`dist_shelf_km`) and krill habitat (`krill_ggp`) continue to show multicollinearity. We will inspect their distribution.  
  
```{r}
#Density distribution for sea ice thickness
den1 <- crab_ind %>% 
  drop_na() %>% 
  ggplot(aes(x = krill_ggp, fill = factor(presence)))+
  geom_density(alpha = 0.5)+
  scale_fill_brewer(palette = "Dark2")+
  theme_bw()

#Density distribution for distance to shelf
den2 <- crab_ind %>% 
  drop_na() %>% 
  ggplot(aes(x = dist_shelf_km, fill = factor(presence)))+
  geom_density(alpha = 0.5)+
  scale_fill_brewer(palette = "Dark2")+
  theme_bw()

#Plotting together
plot_grid(den1, den2, nrow = 2)
```
  
The distributions between presences and background mostly overlap in these two variables, but given that given that krill has a relationship with presence that is significant at 95% probability, we will keep this variable.  
  
## Correlation matrix without distance to continental shelf

```{r}
#Removing freezing potential from num_var data frame and plot correlations again
num_var <- num_var %>% 
  #Removing rows containing NA values
  drop_na() %>% 
  select(!dist_shelf_km)

#Calculating correlations and plotting results
ggcorr(num_var, method = c("pairwise", "spearman"), 
       label = T, label_size = 3, angle = -15)+
  theme(legend.justification = c(1, 0),
        legend.position = c(0.5, 0.7),
        legend.direction = "horizontal") +
  guides(fill = guide_colorbar(barwidth = 9, 
                               barheight = 1, 
                               title.position = "top", 
                               title.hjust = 0.5, 
                               title = "Spearman correlation"))
```
  
We can still see some strong relationships between krill and a few other variables, so will calculate VIF once more.  
  
## Linear regression without distance to continental shelf

```{r}
#Linear regression formula
form_lm <- paste("presence", 
                 paste(colnames(num_var), sep = "+", collapse = "+"), 
                 sep = " ~ ")

#Fitting linear model
model <- lm(as.formula(form_lm), data = crab_ind)

#Checking results
summary(model)
```
  
## Calculating VIF without distance to continental shelf

```{r}
#Calculating VIF values for all predictors in the linear regression
vif_vals <- vif(model)

#Plotting results as a horizontal bar plot
par(mar = c(2, 10, 2, 2))
barplot(vif_vals, horiz = T, las = 1, xlim = c(0, 6))
#We will add a vertical line at 5 to identify variables with large VIF values
abline(v = 5)
```
  
Krill has strong correlations with distance to the sea ice edge (`dist_ice_edge_km`), long-term presence of sea ice (`lt_pack_ice`), and sea ice thickness (`SIT_m`). The relationships between all these variables and presence is significant, but we will check `dist_ice_edge_km` because it significance is at a lower alpha (0.95 instead of <0.001).  
  
## Correlation matrix without distance to sea ice edge

```{r}
#Removing freezing potential from num_var data frame and plot correlations again
num_var <- num_var %>% 
  #Removing rows containing NA values
  drop_na() %>% 
  select(!dist_ice_edge_km)

#Calculating correlations and plotting results
ggcorr(num_var, method = c("pairwise", "spearman"), 
       label = T, label_size = 3, angle = -15)+
  theme(legend.justification = c(1, 0),
        legend.position = c(0.5, 0.7),
        legend.direction = "horizontal") +
  guides(fill = guide_colorbar(barwidth = 9, 
                               barheight = 1, 
                               title.position = "top", 
                               title.hjust = 0.5, 
                               title = "Spearman correlation"))
```
  
## Linear regression without distance to sea ice edge

```{r}
#Linear regression formula
form_lm <- paste("presence", 
                 paste(colnames(num_var), sep = "+", collapse = "+"), 
                 sep = " ~ ")

#Fitting linear model
model <- lm(as.formula(form_lm), data = crab_ind)

#Checking results
summary(model)
```
  
## Calculating VIF without distance to sea ice edge

```{r}
#Calculating VIF values for all predictors in the linear regression
vif_vals <- vif(model)

#Plotting results as a horizontal bar plot
par(mar = c(2, 10, 2, 2))
barplot(vif_vals, horiz = T, las = 1, xlim = c(0, 6))
#We will add a vertical line at 5 to identify variables with large VIF values
abline(v = 5)
```
  
Multicollinearity is still present, but now the significance of relationship between `lt_pack_ice` and presence is at a lower alpha. We will remove this variable and check once more.  
  
## Correlation matrix without long-term presence of pack ice

```{r}
#Removing freezing potential from num_var data frame and plot correlations again
num_var <- num_var %>% 
  #Removing rows containing NA values
  drop_na() %>% 
  select(!lt_pack_ice)

#Calculating correlations and plotting results
ggcorr(num_var, method = c("pairwise", "spearman"), 
       label = T, label_size = 3, angle = -15)+
  theme(legend.justification = c(1, 0),
        legend.position = c(0.5, 0.7),
        legend.direction = "horizontal") +
  guides(fill = guide_colorbar(barwidth = 9, 
                               barheight = 1, 
                               title.position = "top", 
                               title.hjust = 0.5, 
                               title = "Spearman correlation"))
```

## Linear regression without long-term presence of pack ice

```{r}
#Linear regression formula
form_lm <- paste("presence", 
                 paste(colnames(num_var), sep = "+", collapse = "+"), 
                 sep = " ~ ")

#Fitting linear model
model <- lm(as.formula(form_lm), data = crab_ind)

#Checking results
summary(model)
```
  
## Calculating VIF without long-term presence of pack ice

```{r}
#Calculating VIF values for all predictors in the linear regression
vif_vals <- vif(model)

#Plotting results as a horizontal bar plot
par(mar = c(2, 10, 2, 2))
barplot(vif_vals, horiz = T, las = 1, xlim = c(0, 6))
#We will add a vertical line at 5 to identify variables with large VIF values
abline(v = 5)
```
  
All VIF values are under 5, which suggests that multicollinearity between predictor variables is not an issue anymore. We have reduced our predictor variables from 18 to `r ncol(num_var)`. These are the final set of predictor variables that we will use to estimate the distribution of crabeater seals.  
  
## Saving new dataset
This dataset will not contain the variables that were identified as having high correlation with one or more predictor variables.  

```{r save_data, eval = F}
crab_ind %>% 
  #Keep columns containing dates, coords, etc, and low VIF variables
  select(year:presence | all_of(names(num_var))) %>% 
  write_csv("../../Environmental_Data/model_env_pres_bg_20x_Indian_weaning_LowVIF.csv")
```


