---
title: "Exploratory Data Analysis - Environmental variable selection (ACCESS subset)"
author: "Denisse Fierro Arcos"
date: "2023-08-25"
output:
  github_document:
    toc: true
    html_preview: false
---

# Environmental variable selection

Before developing a Generalised Additive Model (GAM) to estimate the distribution of crabeater seals in East Antarctica, we will check all that the 17 environmental variables identified as potential predictors of crabeater seal distribution are not highly correlated with each other. This is known as multicollinearity and it can cause problems with regression analyses as highly correlated variables do not provide independent information to the regression model and this can cause problems when fitting the model.  
  
To identify multicollinearity, we will calculate the correlation among all environmental variables. High correlation (r > 0.75) can be an indicator of multicollinearity. We will follow this by applying a linear regression using all environmental variables as predictors and the `presence` column as the response (a value of `1` in this columns indicates an observation while `0` refers to background data). We will use these results to calculte the variance inflation factor (VIF). The VIF is the most common way to detect multicollinearity with values above 5 indicating a strong correlation across variables.  
  
Finally, to help us decide which environmental variables we will keep if multicollinearity is detected, we will use the results from the linear regression. We will also plot the density distribution of the environmental variables between observation and background points to help identify which variable to keep as a predictor. Variables that appear to show a larger difference in their density distribution will be preferred as predictors as they capture differences in environmental conditions detected at seal sighting location in comparison to background points.  

## Loading libraries

```{r libraries, results = "hide", message = F, warning = F}
library(tidyverse)
library(GGally)
library(car)
```


## Loading observations and background data

We will load these two datasets and merge them together before running our exploratory analysis.

```{r data}
#Getting names of environmental variables available through observations
obs_env <- read_csv("../../Environmental_Data/Env_obs/unique_crabeater_obs_all_env.csv") %>% 
  names()

#Getting environmental data from ACCESS-OM2-01 model
crab_ind <- read_csv("../../Environmental_Data/ACCESS-OM2-01/unique_crabeater_obs_all_env.csv") %>% 
  #Selecting observations for the Indian sector during the weaning period
  filter(str_detect(sector, "Indian") & life_stage == "weaning") %>% 
  #Binding to background data
  bind_rows(read_csv("../../Environmental_Data/ACCESS-OM2-01/unique_background_20x_obs_all_env.csv"))

#Keeping only variables for which observations are available 
crab_ind <- crab_ind[obs_env] %>% 
  #Removing duplicate date information
  select(!c(event_date:date))
  

#Checking data structure
glimpse(crab_ind)
```

## Extracting environmental variables
We will create a new variable to facilitate this exploratory analysis.

```{r}
#Checking correlation between numeric variables
num_var <- crab_ind %>% 
  select(bottom_slope_deg:dist_ice_edge_km) %>% 
  #Removing rows containing NA values
  drop_na()
```


### Calculating and plotting correlation matrix

We will calculate the Spearman correlations across all variables. We use Spearman because it does not assume relationships across variables are linear.

```{r corrplot}
#Calculating correlations and plotting results
ggcorr(num_var, method = c("pairwise", "spearman"), 
       label = T, label_size = 3, angle = -15)+
  theme(legend.justification = c(1, 0),
        legend.position = c(0.5, 0.7),
        legend.direction = "horizontal") +
  guides(fill = guide_colorbar(barwidth = 9, 
                               barheight = 1, 
                               title.position = "top", 
                               title.hjust = 0.5, 
                               title = "Spearman correlation"))
```
There are a number of variables with a strong correlation (0.75 or above). This suggests that multicollinearity would be an issue in our model if we were to include all variables as predictors. We will continue the analysis by calculating the variance inflation factor (VIF). This will help us identify the variables that we should consider excluding from the model due to their lack of independence with one or more predictor variables.


## Calculating linear regression between predicting and response variable
The response variable will be the `presence` column, which identifies observations and background data. The predicting variables are all the variables shown in the correlation plot above.

```{r lm_prep}
# First we will define the formula to be used in the linear regression. We will use the names of all variables in the num_var data frame
form_lm <- paste("presence", 
                 paste(colnames(num_var), sep = "+", collapse = "+"), 
                 sep = " ~ ")

#Fitting linear model
model <- lm(as.formula(form_lm), data = crab_ind)

#Checking results
summary(model)
```

We can see that not all variables under consideration were found to have a significant correlation to the `presence` column. We can now calculate VIF values for all predicting variables.
  
## Calculating and plotting VIF values

```{r vif}
#Calculating VIF values for all predictors in the linear regression
vif_vals <- vif(model)

#Plotting results as a horizontal bar plot
par(mar = c(2, 10, 2, 2))
barplot(vif_vals, horiz = T, las = 1)
#We will add a vertical line at 5 to identify variables with large VIF values
abline(v = 5)
```

Sea ice concentration (`SIC`), distance to the continental shelf (`dist_shelf_km`) and sea surface temperature (`SST_degC`) are the three factors with the largest VIF values. Since SIC was found to be NOT significant in the model, we will remove it from further analyses. We will check for multicollinearity once removed.  
  
## Correlation matrix without sea ice concentration

```{r}
#Removing freezing potential from num_var data frame and plot correlations again
num_var <- num_var %>% 
  #Removing rows containing NA values
  drop_na() %>% 
  select(!SIC)

#Calculating correlations and plotting results
ggcorr(num_var, method = c("pairwise", "spearman"), 
       label = T, label_size = 3, angle = -15)+
  theme(legend.justification = c(1, 0),
        legend.position = c(0.5, 0.7),
        legend.direction = "horizontal") +
  guides(fill = guide_colorbar(barwidth = 9, 
                               barheight = 1, 
                               title.position = "top", 
                               title.hjust = 0.5, 
                               title = "Spearman correlation"))
```
  
## Linear regression without SIC
High correlations are still detected, so we will run the linear regression and calculate VIF once again.
  
```{r}
#Linear regression formula
form_lm <- paste("presence", 
                 paste(colnames(num_var), sep = "+", collapse = "+"), 
                 sep = " ~ ")

#Fitting linear model
model <- lm(as.formula(form_lm), data = crab_ind)

#Checking results
summary(model)
```

## Calculating VIF without SST

```{r}
#Calculating VIF values for all predictors in the linear regression
vif_vals <- vif(model)

#Plotting results as a horizontal bar plot
par(mar = c(2, 10, 2, 2))
barplot(vif_vals, horiz = T, las = 1)
#We will add a vertical line at 5 to easily identify variables with large VIF values
abline(v = 5)
```
  
Distance to the shelf is highly correlated to depth, and both these variables have relationships to presence that are significant. However, given that the significance of depth is at a lower alpha, we will remove distance to the shelf and test for multicollinearity once more.  
  
## Correlation matrix without distance to shelf

```{r}
#Removing freezing potential from num_var data frame and plot correlations again
num_var <- num_var %>% 
  #Removing rows containing NA values
  drop_na() %>% 
  select(!dist_shelf_km)

#Calculating correlations and plotting results
ggcorr(num_var, method = c("pairwise", "spearman"), 
       label = T, label_size = 3, angle = -15)+
  theme(legend.justification = c(1, 0),
        legend.position = c(0.5, 0.7),
        legend.direction = "horizontal") +
  guides(fill = guide_colorbar(barwidth = 9, 
                               barheight = 1, 
                               title.position = "top", 
                               title.hjust = 0.5, 
                               title = "Spearman correlation"))
```
  
There are still some variables that are highly correlated, so we will check VIF again.  
  
## Linear regression without distance to shelf

```{r}
#Linear regression formula
form_lm <- paste("presence", 
                 paste(colnames(num_var), sep = "+", collapse = "+"), 
                 sep = " ~ ")

#Fitting linear model
model <- lm(as.formula(form_lm), data = crab_ind)

#Checking results
summary(model)
```
  
## Calculating VIF without distance to shelf

```{r}
#Calculating VIF values for all predictors in the linear regression
vif_vals <- vif(model)

#Plotting results as a horizontal bar plot
par(mar = c(2, 10, 2, 2))
barplot(vif_vals, horiz = T, las = 1, xlim = c(0, 6))
#We will add a vertical line at 5 to easily identify variables with large VIF values
abline(v = 5)
```
  
All VIF values are under 5, which suggests that multicollinearity across predictor variables is not an issue anymore. We have reduced our predictor variables from 8 to 7. These are the final set of predictor variables that we will use to estimate the distribution of crabeater seals.  
  
## Saving new dataset
This dataset will not contain the variables that were identified as having high correlation with one or more predictor variables.  

```{r save_data, eval = F}
crab_ind %>% 
  #Keep columns containing dates, coords, etc, and low VIF variables
  select(year:presence | all_of(names(num_var))) %>% 
  write_csv("../../Environmental_Data/ACCESS-OM2-01/Obs_BG_20x_Indian_weaning_LowVIF_match-obs.csv")
```


