---
title: "Exploratory Data Analysis - Environmental variable selection"
author: "Denisse Fierro Arcos"
date: "2023-07-24"
output:
  github_document:
    toc: true
    html_preview: false
---

# Environmental variable selection

Before developing a Generalised Additive Model (GAM) to estimate the distribution of crabeater seals in East Antarctica, we will check all that the 17 environmental variables identified as potential predictors of crabeater seal distribution are not highly correlated with each other. This is known as multicollinearity and it can cause problems with regression analyses as highly correlated variables do not provide independent information to the regression model and this can cause problems when fitting the model.  
  
To identify multicollinearity, we will calculate the correlation among all environmental variables. High correlation (r > 0.75) can be an indicator of multicollinearity. We will follow this by applying a linear regression using all environmental variables as predictors and the `presence` column as the response (a value of `1` in this columns indicates an observation while `0` refers to background data). We will use these results to calculte the variance inflation factor (VIF). The VIF is the most common way to detect multicollinearity with values above 5 indicating a strong correlation across variables.  
  
Finally, to help us decide which environmental variables we will keep if multicollinearity is detected, we will use the results from the linear regression. We will also plot the density distribution of the environmental variables between observation and background points to help identify which variable to keep as a predictor. Variables that appear to show a larger difference in their density distribution will be preferred as predictors as they capture differences in environmental conditions detected at seal sighting location in comparison to background points.  

## Loading libraries

```{r libraries, results = "hide", message = F, warning = F}
library(tidyverse)
library(GGally)
library(car)
```


## Loading observations and background data

We will load these two datasets and merge them together before running our exploratory analysis.

```{r data}
crab_ind <- read_csv("../../Environmental_Data/Env_obs/unique_crabeater_obs_all_env.csv") %>%  
  #Selecting observations for the Indian sector during the weaning period
  filter(str_detect(sector, "Indian") & life_stage == "weaning") %>% 
  #Removing original date column
  select(!event_date) %>% 
  #Binding to background data
  bind_rows(read_csv("../../Environmental_Data/Env_obs/unique_background_20x_obs_all_env.csv"))

#Checking data structure
glimpse(crab_ind)
```

## Extracting environmental variables
We will create a new variable to facilitate this exploratory analysis.

```{r}
#Checking correlation between numeric variables
num_var <- crab_ind %>% 
  select(bottom_slope_deg:dist_ice_edge_km) %>% 
  #Removing rows containing NA values
  drop_na()
```


### Calculating and plotting correlation matrix

We will calculate the Spearman correlations across all variables. We use Spearman because it does not assume relationships across variables are linear.

```{r corrplot}
#Calculating correlations and plotting results
ggcorr(num_var, method = c("pairwise", "spearman"), 
       label = T, label_size = 3, angle = -15)+
  theme(legend.justification = c(1, 0),
        legend.position = c(0.5, 0.7),
        legend.direction = "horizontal") +
  guides(fill = guide_colorbar(barwidth = 9, 
                               barheight = 1, 
                               title.position = "top", 
                               title.hjust = 0.5, 
                               title = "Spearman correlation"))
```
There are a number of variables with a strong correlation (0.75 or above). This suggests that multicollinearity would be an issue in our model if we were to include all variables as predictors. We will continue the analysis by calculating the variance inflation factor (VIF). This will help us identify the variables that we should consider excluding from the model due to their lack of independence with one or more predictor variables.


## Calculating linear regression between predicting and response variable
The response variable will be the `presence` column, which identifies observations and background data. The predicting variables are all the variables shown in the correlation plot above.

```{r lm_prep}
# First we will define the formula to be used in the linear regression. 
#We will use the names of all variables in the num_var data frame
form_lm <- paste("presence", 
                 paste(colnames(num_var), sep = "+", collapse = "+"),
                 sep = " ~ ")

#Fitting linear model
model <- lm(as.formula(form_lm), data = crab_ind)

#Checking results
summary(model)
```
  
We can see that not all variables under consideration were found to have a significant correlation to the `presence` column. We can now calculate VIF values for all predicting variables.
  
## Calculating and plotting VIF values

```{r vif}
#Calculating VIF values for all predictors in the linear regression
vif_vals <- vif(model)

#Plotting results as a horizontal bar plot
par(mar = c(2, 10, 2, 2))
barplot(vif_vals, horiz = T, las = 1, xlim = c(0, 6))
#We will add a vertical line at 5 to identify variables with large VIF values
abline(v = 5)
```
  
Distance to the continental shelf (`dist_shelf_km`) is the only factors with a VIF large enough to indicate multicollinearity. This variable is highly correlated to depth. Since `dist_shelf_km` was found to be NOT significant in the model, we will remove it from further analyses. We will check for multicollinearity once removed.  
  
## Correlation matrix without distance to shelf

```{r}
#Removing freezing potential from num_var data frame and plot correlations again
num_var <- num_var %>% 
  #Removing rows containing NA values
  drop_na() %>% 
  select(!dist_shelf_km)

#Calculating correlations and plotting results
ggcorr(num_var, method = c("pairwise", "spearman"), 
       label = T, label_size = 3, angle = -15)+
  theme(legend.justification = c(1, 0),
        legend.position = c(0.5, 0.7),
        legend.direction = "horizontal") +
  guides(fill = guide_colorbar(barwidth = 9, 
                               barheight = 1, 
                               title.position = "top", 
                               title.hjust = 0.5, 
                               title = "Spearman correlation"))
```
  
Some variables still have some high correlation values, so we will calculate VIF values.  
  
## Linear regression without distance to shelf
  
```{r}
#Linear regression formula
form_lm <- paste("presence", 
                 paste(colnames(num_var), sep = "+", collapse = "+"), 
                 sep = " ~ ")

#Fitting linear model
model <- lm(as.formula(form_lm), data = crab_ind)

#Checking results
summary(model)
```
  
## Calculating VIF without distance to shelf

```{r}
#Calculating VIF values for all predictors in the linear regression
vif_vals <- vif(model)

#Plotting results as a horizontal bar plot
par(mar = c(2, 10, 2, 2))
barplot(vif_vals, horiz = T, las = 1, xlim = c(0, 6))
#We will add a vertical line at 5 to easily identify variables with large VIF values
abline(v = 5)
```

## Saving new dataset
This dataset does not contain any more variables that are highly correlated, so we will save a copy of the results.  

```{r save_data, eval = F}
crab_ind %>%
  #Keep columns containing dates, coords, etc, and low VIF variables
  select(year:presence | all_of(names(num_var))) %>% 
  write_csv("../../Environmental_Data/Env_obs/Obs_BG_20x_Indian_weaning_LowVIF.csv")
```


